---
title: "Smith et al Manuscript"
author: "Steven Smith"
date: "November 6, 2017"
output:
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---
# Analysis Script
**The following markdown serves as a record for analysis performed in Smith et al., 2017.**   
**Change the root directory to point to where "Script_input" and "Scripts" directories reside. This is currently hardcoded as the Git repo **  
This script will automatically create the following directories, if not present already:  
1.  Tables- Tables associated with manuscript, in .csv or .txt  
2.  Figures - Figures associated with manuscript, in .eps  
3.  Script_output - anything else output that is not a figure or table   

A sessionsInfo log file will be included in Script_output.  

Script_input contains static data used by script (previously generated and formatted for use as input). Also includes a data dictionary for sample metadata. 

Some library installation may be required (e.g., install.packages("PACKAGE"))

** Some packages may mask fucntions from other packages. Not sure how to remedy this, except for explicilty calling all functions with the package name. 
** cairo_ps may not function proportly on macOS systems without XQuartz (https://www.xquartz.org). Either download Xquartz or subsititue cairo_ps for postscript(). Note that postscript() is not able to print transparency. 


Need to followup

-Figure 2_2_X
-Ref Gvag vs Media-- RNAseq vs qPCR
-GO figure
-Figure_3_4 and 5?
-Sctaych area vs EdU
-Long plots- only the Lacto RA**


```{r timestamp1}
timestamp()
# Clear enviornment variables, set root 
rm(list=ls())

## Set the root directory here:
root_directory<-"~/smith_thesis_2017/AnalysisPipeline/"#"~/Dropbox (IGS)/Jacques_Steve_Shared/Thesis/Thesis_pipeline/AnalysisPipeline/"

## Supress warnings to make knit PDF shorter... but turn these back on since there may be some weird behaviors
knitr::opts_chunk$set(warning=FALSE, message=T,size=8)

```
# Prepare Enviornment    

1. Setup environment, variables, etc   
2. Define custom functions   
```{r setup,eval=TRUE,warning=FALSE}
## Load libraries 
## Note that some packages mask others. This might be a problem for, e.g., rename. 

## Load libraries.
## Note that some override/mask functions from others. Had to explicitly use the "dplyr" package for all "select" statements
library(reshape)
library(ggbiplot)
library(plyr)
library(grid)
library(scales)
library(gridExtra)
library(edgeR) 
library(RColorBrewer)
library(psych)
library(randomForest) 
library(rfPermute)
library(tidyverse)
library(squash)
library(stringr)
library(plotly)
library(gPCA)
library(nlme)
library(gplots)
library(Biobase)
library(caret)

## Source the Random Forest wrapper script
source(paste0(root_directory,"Scripts/rfSubjectSpecific.R"))

## Set output directories for tables, figures and data structures. 
if(!dir.exists(file.path(root_directory, "Figures"))){
  dir.create(file.path(root_directory, "Figures"))  
}
if(!dir.exists(file.path(root_directory, "Tables"))){
  dir.create(file.path(root_directory, "Tables"))  
}
if(!dir.exists(file.path(root_directory, "Script_output"))){
  dir.create(file.path(root_directory, "Script_output"))  
}

R_script_output_directory<-paste0(root_directory,"Script_output/")
R_script_input_directory<-paste0(root_directory,"Script_input/") ## Contains Rdata files created outside of this script. These data are static. 
thesis_tables_directory<-paste0(root_directory,"Tables/")
thesis_figures_directory<-paste0(root_directory,"Figures/")

##Set global variables
seed_val<-4543 ## Needed for exact results obtained in manuscript
pval_threshold<-0.05 ## Signifgance value threshold
npermutes<-500 ## Number of permutations to generate emperical null distribution in RF models
nfolds<-10 #Number of cross-fold validation in customized RF script
training_prop<-0.7 ## Proportion of input data for use in training models
nSpecies<-9 ## Max # of species to plot in Fig 1
sizes<-1 ## Default sizes for plots
raThreshold <-0.02 ## Min relative abundance threshold to plot in Fig 1
margins<-unit(c(-2.5,40,-2.5,5),units="points") ## Default margins for plots
ph_normalization_factor<-3 ## rescales y axis so that plot doesn't bein at 0
alpha_rect<-0.7 ## Fig 1 greyed out rectangle opacity
rect_fill<-"grey" ## Fig 1 greyed out color
removed_samples<-data.frame(Pre_QC_ID=NULL,QC_removal_stage=NULL)## Keep a list of removed samples, and the stage of removal

## Set global plot theme
journalFormat <- theme_bw() + theme(text = element_text(family = "Arial", colour = "black",size=12))

## Set standardized color table
load(file=paste0(R_script_input_directory,"subject_long_taxa_colors.Rdata"))
color_scheme_BCS<- c("L. crispatus"= unname(subject_long_taxa_colors["Lactobacillus_crispatus"]),
                     "L. jensenii" = unname(subject_long_taxa_colors["Lactobacillus_jensenii"]),
                     "L. iners" = unname(subject_long_taxa_colors["Lactobacillus_iners"]),
                     "G. vaginalis" = unname(subject_long_taxa_colors["Gardnerella_vaginalis"]),
                     "Cell Culture Medium"='blue',
                     "0.1% D-lactic acid"="#4dac26",
                     "0.1% L-lactic acid"="#d01c8b",
                     "1% lactic acid, pH 7.66"="#386cb0",
                     "0.06% L-lactic acid"="#f1b6da",
                     "0.06% D-lactic acid"="#b8e186",
                    # "0.1% DL-lactic acid"= "#b6daf1"
                     "DL_10"= "#b6daf1",
                    "0.1% DL lactic acid" = "#b6daf1",
                    "0.1% DL-lactic acid" = "#b6daf1"
) 

cst.colors<-c("I-A"="red1",
              "I-B"="#990000",
              "I" = unname(subject_long_taxa_colors["Lactobacillus_crispatus"]),
              "II"=unname(subject_long_taxa_colors["Lactobacillus_gasseri"]),
              "III-A"="darkorange" ,
              "III-B"="#cc7a00" ,
              "III"=unname(subject_long_taxa_colors["Lactobacillus_iners"]),
              "IV-A"="lightseagreen",
              "IV-B"="mediumblue",
              "V"=unname(subject_long_taxa_colors["Lactobacillus_jensenii"]),
              "DUMMY"='grey'
)


## Table and figure names
TABLE_SEQSUMMARY<-"TABLE_1.csv"
TABLE_TOPMIRS<-"TABLE_2.txt"
TABLE_TRL_NUMDEGENES<-"TABLE_3.csv"
TABLE_TRL_SUMMARY_PATHWAYS<-"TABLE_4.csv"

TABLE_COUNTS_RAW<-"TABLE_S1.csv"
TABLE_MODEL_INPUT<-"TABLE_S2.csv"
TABLE_SRL_METADATA<-"TABLE_S3.csv"
#TABLE_PROXY_AMSEL_INPUT<-"TABLE_S4.csv"
#TABLE_RF_SUMMARY.CV<-"TABLE_S5.csv"
#TABLE_PROXY_AMSEL_SRL<-"TABLE_S6.csv"
TABLE_RF_SUMMARY<-"TABLE_S7.csv"
TABLE_MIR_TARGETS<-"TABLE_S8.txt"
TABLE_QPCR_TIMECOURSE<-"TABLE_S9.csv"
TABLE_EDU_SCRATCH_QUANT<-"TABLE_S10.csv"
TABLE_CT_QUANT<-"TABLE_S11.csv"
TABLE_TRL_ALIGNSTATS<-"TABLE_S12.csv"
TABLE_TRL_COUNTS_RAW<-"TABLE_S13.csv"
TABLE_EDGER_RESULTS<-"TABLE_S14.csv"
TABLE_TRL_PATHWAY_Z_SCORES<-"TABLE_S15.csv"

FIGURE_SUBJECT_PLOTS<-"FIGURE_1_"
FIGURE_RF_IMPORTANCE_AMSEL<-"FIGURE_drop_2.eps"
FIGURE_RF_IMPORTANCE<-"Figure_drop.eps"
FIGURE_TOPMIRS<-"FIGURE_2.eps"
FIGURE_QPCR_TIMECOURSE<-"FIGURE_3.eps" ## Add lactic acid or put as supp;
FIGURE_EDU_QUANT<-"FIGURE_4_D.ps"
FIGURE_SCRATCH_QUANT<-"FIGURE_4_B.ps"
FIGURE_CT_INFECT_QUANT<-"Figure_6A.ps"
FIGURE_CT_EDU_QUANT<-"Figure_6B.ps"
FIGURE_WESTERN_QUANT<-"Figure_5B.ps"

FIGURE_COMBINED_PATHWAYS_CYCLE<-"Figure_7.eps"
FIGURE_LONGITDUINAL_GENEEXP.cycle<-"Figure_8.eps"
FIGURE_FINAL_MODEL<-"Figure_9.eps" ##Placeholder

FIGURE_QC<-"FIGURE_S1_"
FIGURE_QC_RIN_v_READS.PROP<-paste0(FIGURE_QC,"1.eps")
FIGURE_QC_RIN_v_READS.ABS<-paste0(FIGURE_QC,"2.eps")
FIGURE_QC_PCA.PREQC.BYBATCH<-paste0(FIGURE_QC,"3.eps")
FIGURE_QC_PCA.PREQC.BYSUBJ<-paste0(FIGURE_QC,"4.eps")
FIGURE_QC_PCA.RMLOW.BYBATCH<-paste0(FIGURE_QC,"5.eps")
FIGURE_QC_PCA.RMLOW.BYSUBJ<-paste0(FIGURE_QC,"6.eps")
FIGURE_QC_PCA.NORMAL.BYBATCH<-paste0(FIGURE_QC,"7.eps")
FIGURE_QC_PCA.NORMAL.BYSUBJ<-paste0(FIGURE_QC,"8.eps")
FIGURE_MIR_TARGETS_GO<-"Figure_S2.eps"
FIGURE_DL_LACTICACID_QPCR<-"Figure_S3.eps" ## maybe add lactic acid time course here and split out 13h D, L, DL panel. 
##S4 =  EdU microscopy
FIGURE_nyc_v_tsb<-"Figure_S5.eps" ## Add EdU barplot as well. 
#FIGURE_TRL_RIN_HIST<-"Figure_.eps"
#FIGURE_COMBINED_PATHWAYS_IMMUNE<-"Figure_S7.eps"
#FIGURE_LONGITDUINAL_GENEEXP.immune<-"FIGURE_S9.eps"

setwd(paste0(R_script_output_directory))


```

# Custom functions

```{r custom_functions,eval=TRUE}
###########
## remove_poorQC_samples function
###########

## Adds samples in sample_list to running removed_samples data.frame. Tracks the reaon for removal
remove_poorQC_samples<-function(removed_samples=removed_samples,
                                sample_list=c(""),
                                reason=""){
  
  removed_samples<-unique(rbind(removed_samples,data.frame(Pre_QC_ID=sample_list,QC_removal_stage=reason)))
  return(removed_samples)
}

###########
## End remove_poorQC_samples function
###########


###########
## plot_pca function
###########

## Plots PCA and calcualtes gPCA-based p value. Outputs a list with PCA plot object and p value (if plotly=F). Otherwise, just a plotly plot object 

plot_pca<-function(ES, ## Expression Set object
                   plot_title="", ## Title for PCA plot
                   center=TRUE, ## pcrcomp centering?
                   scale=TRUE, ##prcomp scaling?
                   color_by='SID', ## What to color plots by
                   logt=TRUE, #Whether counts in ES should be log transformed
                   obs.scale = 1, ##prcomp scale factor
                   var.scale = 1, ##prcomp var scaling
                   ellipse = FALSE, ## draw ellipse around groups in PCA plot
                   circle = FALSE, ## or circle
                   var.axes=FALSE, ##option for prcomp
                   ploly=FALSE, ##whether to generate a ploty interactive plot. This option will not return gPCA p value
                   margins=unit(c(0,0,0,0),units = "points"),
                   seed_val=4543, ## seed needed for gPCA
                   ...){ 
  
  ## es = ExpressionSet object  
  ## ** Assumes ES counts data is log transformed. Set logt=FALSE to logt data first. 
  ## **Assumes columns named "Batch", "SID" ,"BVGroup" and "NUGENT_CLASS" exist in pData in ES. 
  
  set.seed(seed_val) ## Repeatable results in Guided PCA
  ##calcualte variance of counts
  variable_counts<-t(exprs(ES))[,apply(t(exprs(ES)), 2, var, na.rm=TRUE) != 0]
  
  ## log transform data if needed
  if(logt){
    cnts <- log(variable_counts+1,base = 2)
  }else{
    cnts<-variable_counts
  }
  
  ## extract metadata from ES to decorate tree. Assumes these variables are present
  batch <- as.numeric(pData(ES)[, 'Batch']) 
  sid<-pData(ES)[, 'SID']
  bvgroup<-pData(ES)[,'BVGroup']
  bvclass<-pData(ES)[,'NUGENT_CLASS']
  ##color mappings for different metadata
  if(color_by=="SID"){
    group<-as.character(sid)
    group.gpca<-as.numeric(as.factor(group))
  }else if(color_by=="Batch"){
    group<-as.character(batch)
    group.gpca<-batch
  }else if (color_by=="NugentC"){
    group<-as.character(bvclass)
    group.gpca<-as.numeric(as.factor(group))
  }
  else{
    group<-as.character(bvgroup)
    group.gpca<-as.numeric(as.factor(group))
  }
  
  ##PCA
  counts.pca <- prcomp(cnts,center = center, scale. = scale) 
  
  #PCA plot
  pca.p <- ggbiplot(counts.pca, groups = group, 
                    obs.scale = obs.scale, var.scale = var.scale , 
                    ellipse = ellipse, 
                    circle = circle,var.axes =var.axes, ...)
  pca.p <- pca.p + 
    scale_color_discrete(name = '')+
    journalFormat+
    theme(plot.margin = margins)
  
  gPCA.result<-gPCA.batchdetect(dist(cnts),group.gpca)
  
  ##Optional plotly functionality... returns a list with plot object in an element otherwise (with gPCA p value)
  if(ploly){ggplotly(pca.p)}else{return(list(pca.p=pca.p,gPCA.result=gPCA.result))}
}

###########
## End plot_pca definition
###########

###########
## subset_ExpressionSet function
###########

## Function to subset ExpressionSet objects based on vector of sample/rows names to remove. Returns an expression set without filterout samples

subset_ExpressionSet<-function(expSet, ##expression set
                               filterOut=c(""), ##vector of samples to drop from expSet
                               samples=TRUE){ ## filter samples names or row names
  
  ## Drop sample (columns) from expSet
  if(samples){
    ## Remove count data in filterOut vector
    counts_meta<-exprs(expSet)[,!colnames(exprs(expSet)) %in% filterOut]
    ## Remove metadata in filterOut vector
    design.subset<-pData(expSet)[!row.names(pData(expSet)) %in% filterOut,]
    ## Return a re-packaged filtered count and metadata into ExpressionSet
  }
  else{## Drop miRNAs (rows) from expSet
    counts_meta<-exprs(expSet)[!row.names(exprs(expSet)) %in% filterOut,]
    design.subset<-pData(expSet)[,!colnames(pData(expSet)) %in% filterOut]
    
  }
  return(ExpressionSet(assayData = as.matrix(counts_meta),
                       phenoData = AnnotatedDataFrame(design.subset)))  
}

###########
## End subset_ExpressionSet 
###########

###########
## plot_accuracy 
###########

## Generates predicted vs acutal tables and plots the predicted vs actual values on plot

plot_accuracy<-function(rfp, ## Random Forest object
                        testing_fullset, ## vector of sample names that were held out of training/used for testing
                        index_of_response=match("NUGENT_SCORE",names(testing_fullset)), ## index corresponding to response variable in input data
                        index_of_sid=match("SID",names(testing_fullset)),## index corresponding to subject ID in input data (if subj_spec=T)
                        subj_spec=TRUE, ## whether RF was run with 'subject_spec' option
                        nfold=10) # number of k folds in RF if subj_spec=T
{ 
  
  
  accuracy_table<-data.frame(fold=0,predicted=0,actual=0) ## hold values for actual, predicted values
  
  ## Subject specific rfp have a model for each cross fold, so need to loop through and aggregate each fold
  if(subj_spec){
    
    for(m in 1:nfold){
      #m<-1
      ## Compare predicted to hold out set
      p1<-predict(rfp$mdl[[m]], 
                  testing_fullset[,-c(index_of_response,index_of_sid)], type='response')
      
      accuracy_table<-rbind(accuracy_table,data.frame(fold=m,predicted=p1,actual=testing_fullset[,index_of_response]))
    }
  }else{
    p1<-predict(rfp,testing_fullset[,-c(index_of_response,index_of_sid)], type='response')
    accuracy_table<-data.frame(fold="NA",predicted=p1,actual=testing_fullset[,index_of_response])
  }
  
  accuracy_table<-filter(accuracy_table,!fold==0) ## drop the initialization row
  
  plot_a<-ggplot()+geom_point(data=accuracy_table,aes(y=predicted,x=actual,col=as.factor(fold)))+
    ggtitle("Predicted vs Actual Values from plot_accuracy")+
    journalFormat+
    scale_x_continuous(limits=c(0,10),breaks = 1:10)+
    scale_y_continuous(limits=c(0,10),breaks = 1:10)
  print(plot_a)
  
  return(accuracy_table)
  
}

###########
## End plot_accuracy definition
###########

###########
## run_randomForest definition
###########

## Wrapper to run Random Forest model building. Can run RandomForest, rfPermute or rfSubjectSpecific. Takes care of subsetting data for training/testing and outputting accuracy/error/etc etstimates. Saves/loads previous models. 
#Note if load_prev=T, most of the parameters are ignored
run_randomForest<-function(predictors_response_table,
                           response_variable_name,
                           nfold=10,
                           nreps=105,
                           permute=TRUE,
                           save_model=FALSE,
                           load_prev_model=TRUE,
                           file_n="rf_model",
                           verbose=TRUE,
                           pval_thres=pval_threshold,
                           subj_spec=TRUE,
                           importance_thres=10,
                           training_prop=0.7,
                           seed=seed_val,
                           R_script_output_directory,
                           ...)
{
  
  # predictors_response_table - data frame of predictors + response variable. If it contains a column called 'SID', sets subj_spec=TRUE
  # response_variable_name    - column name as string of response variable (as found in predictors_response_table)
  # nfold                     - number of k-fold validations to run
  # nreps                     - number of permutations to run to compute null distribution. Ignored if permute=FALSE
  # permute                   - whether or not to run rfPermute (generate null distribution permutation p-values for each feature)
  # save_model                - save model as Rdata to outout directory? Uses 'file_n' as file name
  # load_prev_model           - load model outout directory? Uses 'file_n' as file name
  # file_n                    - file name to use when reading or writing a model to disk
  # verbose                   - detailed output of model results, etc
  # pval_thres                -p value threshold to call signifigant 
  # subj_spec                 - whether or not to run rfSubjectSpecific.R. Looks for a column called 'SID' in predictors_response_table
  # importance_thres          -
  #training_prop              - proportion of input data to use as training. remaining is used as hold out set for testing
  # seed                      -seed value to produce repeatable results
  # R_script_output_directory - root output directory for read/write model files
  
  nrep<-nreps
  set.seed(seed)
  ## First major control point: loading from previously saved model, or generating a new trained model? 
  
  #### ///////
  # Start New Model
  #### //////
  
  if(!load_prev_model){
    
    #Find the column index corresponding to response & subject ID
    index_of_response<-match(response_variable_name,names(predictors_response_table))
    index_of_sid<-match("SID",names(predictors_response_table))
    if(is.na(index_of_sid) & subj_spec){
      print("SID could not be found. Setting to non-subject specific")
      subj_spec<-FALSE
      index_of_sid<-0
    }else if (is.na(index_of_sid)){
      index_of_sid<-0
    } else if(index_of_sid>0 & !subj_spec){
      print("Found a SID column. Setting to subject-spefic. ")
      subj_spec<-TRUE
    }
    
    #Partition input data into training and testing
    
    inTrain<-createDataPartition(y=predictors_response_table[,index_of_response],p =training_prop,list = F)
    if(is.character(predictors_response_table[,index_of_response])){
      predictors_response_table[,index_of_response]<-as.factor(predictors_response_table[,index_of_response])
    }
    
    ##Subjet training and testing data
    training_fullset<-predictors_response_table[inTrain,]
    testing_fullset<-predictors_response_table[-inTrain,]
    table(predictors_response_table[,index_of_response])
    table(training_fullset[,index_of_response])
    table(testing_fullset[,index_of_response])
    response<-training_fullset[,index_of_response]
    
    # Determine whether to run classification or regression depending on response variable type
    if(is.numeric(predictors_response_table[,index_of_response])){# regression
      rf_type<-"regression"
    }else{
      rf_type<-"classification"
    }
    
    #Determine whether to run rfSubjectSpecific.R
    if(subj_spec){
      print(paste0("Starting subject-specific rfSubjectSpecific with permute set to : ",permute))
      rfp<-rfSubjectSpecific(training_fullset[,-c(index_of_response,index_of_sid)],response,subjID = as.character(training_fullset[,index_of_sid],nrep=nrep),nfolds = nfold,verbose=verbose,nrep = nreps,permute=permute) ## This will be sourced at the setup section. It is an external script.
    }else{
      print("Starting non subject-specific rfPermute")
      rfp<-rfPermute(training_fullset[,-c(index_of_response)],response,nrep = nreps,...)
    }
    
    ## The accuracy, etc output varies depending on whether which combination of RF were run
    
    ## /////
    ## Regression + Subject-Specific
    ## ////
    
    if(rf_type=="regression" & subj_spec){ ## If it's a regression model
      accuracy_table<-plot_accuracy(rfp,testing_fullset = testing_fullset,index_of_response = index_of_response, index_of_sid = index_of_sid,subj_spec = subj_spec,nfold = nfold) ## see above for this function
      
      ## /////
      ## Classification + Non Subject-Specific
      ## ////
      
    }else if (rf_type=="classification" & !subj_spec){
      p1<-predict(rfp, testing_fullset[,-c(index_of_response,index_of_sid)], subj_spec = subj_spec,type='response')
      (accuracy_table<-table(Var1=p1,Var2=testing_fullset[,index_of_response]))
      
      ## /////
      ## Classification + Subject-Specific
      ## ////
      
    }else if (rf_type=="classification" & subj_spec){
      p1<-predict(rfp$mdl, testing_fullset[,-c(index_of_response,index_of_sid)], subj_spec = subj_spec,type='response')
      (accuracy_table<-lapply(p1, function(x) table(x,testing_fullset[,c(index_of_response)])))
      
      ## /////
      ## All else (Regression + Non Subject-Specific)
      ## ////
      
    }else{
      accuracy_table<-NULL
    }
    
    ## Write model to file + accuracy table and training/testing info
    
    if(save_model){
      save(rfp,file=paste0(R_script_output_directory,file_n,".RData"))
      save(accuracy_table,file=paste0(R_script_output_directory,file_n,"_accuracyTable.RData"))
      training_testing<-list(training_fullset=training_fullset,testing_fullset=testing_fullset)
      save(training_testing,file=paste0(R_script_output_directory,file_n,"_training_testing.RData"))
    }
    
    #### ///////
    # Load Previous Model
    #### //////
  }else{
    
    rfp.pointer<-load(file=paste0(R_script_output_directory,file_n,".RData"))
    rfp<-get(rfp.pointer)
    
    accuracy_table.pointer<-load(file=paste0(R_script_output_directory,file_n,"_accuracyTable.RData"))
    accuracy_table<-get(accuracy_table.pointer)
    
    training_testing.pointer<-load(file=paste0(R_script_output_directory,file_n,"_training_testing.RData"))
    training_testing<-get(training_testing.pointer)
    training_fullset<-training_testing$training_fullset
    testing_fullset<-training_testing$testing_fullset
  }
  if(verbose){
    print(rfp)
    print(rfp$mdl)
  }
  
  if(permute & subj_spec){
    importance<-data.frame(rfp$importance) ##only happens if rfSubjectSpecific is run with permute
    importance.pval<-dplyr::select(importance,ends_with("pval"))
  }else{
    if(!subj_spec){
      importance<-data.frame(rp.importance(rfp))
    }else{
      importance<-data.frame(rfp$imp)
    }
  }
  return(list(rfp=rfp, ##model
              accuracy_table=accuracy_table,
              importance=importance,
              training_ids=row.names(training_fullset),
              testing_ids=row.names(testing_fullset))) 
}

###########
## END run_randomForest function
###########


###########
## plot_RIN
###########

## Is there an effect due to RIN and number of reads miRNAs?
### # Wrapper for plot of reads miRNAs vs RIN

plot_RIN_meta<-function(SRL_meta_table, ## Table containing SRL metadata
                        y_series="reads_surviving.percent", ## y axis column name to be plotted
                        col_by="Batch", ## color by
                        vjust = 0,  ## vertical justification
                        nudge_y = 0.05, ## nudge y by...
                        angle = 0, ## y axis angle
                        hjust = 0, ## hortizontal adjustment
                        nudge_x = 0.05, ## nudge x by..
                        check_overlap = FALSE, ## check for points overlap- try to minimize
                        y_series_label=y_series # y label on plot
){
  ## Plot for RIN vs a measure of read mapping
  p.RIN<-ggplot(SRL_meta_table,aes(x=RIN,y=SRL_meta_table[,y_series],col=as.factor(SRL_meta_table[,col_by]),label=SRL_meta_table$Pre_QC_ID),label = Pre_QC_ID)+
    geom_point()+
    ylab(y_series_label)+
    xlab("RINe")+
    # ggtitle(paste(y_series," vs RIN"))+
    geom_text(check_overlap = check_overlap,vjust = vjust, nudge_y = nudge_y,angle = angle,hjust = hjust, nudge_x = nudge_x)+
    journalFormat+
    guides(col=guide_legend(title=col_by))
  return(p.RIN)
}

###########
## end plot_RIN
###########

###########
## mapping_stats
###########

## Generate simple mapping stats given an Expression Set pData object
mapping_stats<-function(column=pData(counts_meta.qc)$number_reads_mirs){
  return(data.frame(min=min(column,na.rm = T),median=median(column,na.rm = T),max=max(column,na.rm = T)))
}

###########
## end mapping_stats
###########


###########
## plot_importance
###########

## Plot RF Importance variables. Output is a plot

plot_importance<-function(importance_df=Lacto_RF$importance, ## Data frame holding importance results
                          ntopfeats=length(Lacto_RF$top_features$top_features.all), ## number of features to plot
                          nfeats=25,#max # of features to plot
                          rankBy="IncMSE", ## Importance variable to rank by
                          model_name="",
                          size_font=12,
                          size_points=5
){
  if(nfeats<ntopfeats){
    stop("Number of top features less than total # of features plotted")
  }
  names(importance_df)<-gsub(pattern = "X\\.",replacement  = "",x = names(importance_df))
  importance_df<-importance_df[order(importance_df[,rankBy],decreasing = T),]
  importance_df.cut<-importance_df[1:nfeats,]
  importance_df.cut$features<-factor(row.names(importance_df.cut),levels = rev(row.names(importance_df.cut)),ordered = T)
  
  importance_df.cut.tmp<-separate(melt(importance_df.cut,id.vars = c("features")),col = variable,into=c("metric","SUFFIX"),sep = "\\.",fill = "right")
  
  importance_df.cut.tmp[is.na(importance_df.cut.tmp$SUFFIX),"SUFFIX"]<-"metric_val"
  importance_df.cut<-spread(data=importance_df.cut.tmp,key = SUFFIX,value = value)
  cutoff<-nfeats+0.5-ntopfeats 
  
  ggplot(importance_df.cut)+
    geom_point(aes(y=features,x=metric_val,col=-log(pval,10),pch=metric),size=size_points)+    
    xlab("Importance Metric")+
    ylab("Predictor")+
    scale_colour_gradient(high="red", low="blue",guide = guide_legend(title = "-Log(p-value)"))+
    journalFormat+
    theme(text=element_text(size = size_font),panel.border = element_rect(size=1))+
    ggtitle(paste(model_name," RF Importance Plot for the First",nfeats,"Ranked Features (by",rankBy,")"))
  
}

###########
## end plot_importance
###########

###########
## t.test2
###########

## A t test using sumamry statistics ( and not the entire sample dataset as with t.test)
## Returns p value, difference of means, standard error and t statistc
t.test2 <- function(m1, ## mean of sample set 1
                    m2, ## mean of sample set 2
                    s1, ## standard dev of sample set 1
                    s2, ## standard dev of sample set 2
                    n1, ## number of samples in sample set 1
                    n2, ## number of samples in sample set 1
                    m0=0, ## the null for hypothesis to test (mean value)=
                    equal.variance=FALSE) #whether to assume equal variance between sample sets
  
{
  if( equal.variance==FALSE ) 
  {
    ## "normalize" standard deviations if unequal variance to compute standard error
    se <- sqrt( (s1^2/n1) + (s2^2/n2) )
    # welch-satterthwaite df
    df <- ( (s1^2/n1 + s2^2/n2)^2 )/( (s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1) )
  } else
  {
    # pooled standard deviation, scaled by the sample sizes
    se <- sqrt( (1/n1 + 1/n2) * ((n1-1)*s1^2 + (n2-1)*s2^2)/(n1+n2-2) )
    
    df <- n1+n2-2
  }      
  t <- (m1-m2-m0)/se## calcualte t statistic
  dat <- c(m1-m2, se, t, 2*pt(-abs(t),df))    ## calculate p value based on Student's t distribution
  names(dat) <- c("Difference of means", "Std Error", "t", "p-value")
  return(dat) 
}

###########
## end t.test2
###########


###########
## apply_ttest
###########

apply_ttest<-function(sigtest_results=proliferation_sigtest,pval_threshold=pval_threshold,summary_stats,test_function="t.test",value="value",obs=""){
  
  ## Loop through each test comparison and apply stat test. Store in dataframe
  for(sigtest in 1:nrow(sigtest_results)){
    print(sigtest)
    #sigtest<-1
    x1<-as.character(sigtest_results[sigtest,"xref"]) 
    y1<-as.character(sigtest_results[sigtest,"reference"])
    
    ## Apply a t test to x1 vs y1
    if(test_function=="t.test"){
      tes<-t.test(x=filter(summary_stats,BCS==x1 )$percent_cells,y=filter(summary_stats,BCS==y1)$percent_cells,alternative = "two.sided") 
      delta_mean<-tes$estimate[2]-tes$estimate[1]
      pval<-tes$p.value
    }else if(test_function=="t.test2"){
      obs<-as.character(sigtest_results[sigtest,"Observation"])
      m1<-filter(summary_stats,Observation==obs & Treatment==x1 )$grand_mean
      m2<-filter(summary_stats,Observation==obs & Treatment==y1 )$grand_mean
      s1<-filter(summary_stats,Observation==obs & Treatment==x1 )$grand_sd
      s2<-filter(summary_stats,Observation==obs & Treatment==y1 )$grand_sd
      n1<-filter(summary_stats,Observation==obs & Treatment==x1 )$n
      n2<-filter(summary_stats,Observation==obs & Treatment==y1 )$n
      print(obs)
      tes<-t.test2(m1 = m1,s1 = s1, m2=m2,s2=s2,n1=n1,n2=n2) #less
      print(tes)
      delta_mean<-tes["Difference of means"]
      pval<-tes["p-value"]
      
      
    }else{
      print(paste0(test_function," not found"))
    }
    
    ## Store p value and difference of means
    sigtest_results[sigtest,"pval"]<-pval
    sigtest_results[sigtest,"mean_diff"]<-delta_mean
  }
  
  ## Denote signifigant tests with "*"
  sigtest_results[sigtest_results$pval<=pval_threshold,"sig"]<-"*"
  return(sigtest_results)
}

###########
## end apply ttest
###########

###########
## setup_sigtest
###########
#pval_threshold = pval_threshold;raw_data = proliferation.m;test_function = "t.test";Experiment = "Scratch"

setup_sigtest<-function(pval_threshold = pval_threshold,raw_data = raw_data,test_function = "t.test",Experiment="Scratch"){
  ## Set up a significance test df to store inference testing data
  if(Experiment=="Scratch"){
    sigtest<-data.frame(xref=c(rep("L. crispatus",times=4),rep("L. jensenii",times=3),rep("L. iners",times=2),"G. vaginalis"),reference=c("L. jensenii","L. iners","G. vaginalis","Cell Culture Medium","L. iners","G. vaginalis","Cell Culture Medium","G. vaginalis","Cell Culture Medium","Cell Culture Medium"))
    
    ## Factor labels
    sigtest$xref<-factor(sigtest$xref,levels =c("L. crispatus","L. jensenii","L. iners","G. vaginalis","Cell Culture Medium"),ordered = T)
    sigtest$reference<-factor(sigtest$reference,levels = c("L. jensenii","L. iners","G. vaginalis","Cell Culture Medium"),ordered = T)
    
    sigtest<-apply_ttest(sigtest_results = sigtest,pval_threshold = pval_threshold,summary_stats = raw_data,test_function = "t.test")
    ## Summarize scratch assay data with mean and sd
    summary_stats<-ddply(dplyr::select(raw_data,c(BCS,percent_cells)),c("BCS"),summarise,mean=mean(percent_cells),sd=sd(percent_cells))
    
  }else if(Experiment=="Infection"){
    
    sigtest<-data.frame(Observation=rep(c("Proliferation","Infectivity"),each=3),xref=c(rep("CAS 546102-60-7",times=2),rep("Fascaplysin",times=1)),reference=c("Fascaplysin","Cell Culture Medium","Cell Culture Medium"))
    
    sigtest$xref<-factor(sigtest$xref,ordered = T,levels = c("CAS 546102-60-7","Fascaplysin"))
    sigtest$reference<-factor(sigtest$reference,ordered = T,levels = c("Fascaplysin","Cell Culture Medium"))
    
    
    raw_data.spread<-spread(raw_data,key = Observation,value = percent_cells)
    plot(raw_data.spread$Infectivity,raw_data.spread$Proliferation,col=raw_data.spread$Treatment)
    
    summary_stats.pre<-ddply(raw_data,c("Observation","Treatment","Coverslip"),summarise,mean=mean(percent_cells,na.rm = T),sd=sd(percent_cells,na.rm = T))
    
    summary_stats<-ddply(summary_stats.pre,c("Treatment","Observation"),summarise,grand_mean=mean(mean,na.rm = T))
    
    for(obs in c("Infectivity","Proliferation")){
      for(treat in unique(summary_stats$Treatment)){
        #treat<-"Control"
        gm<-summary_stats[summary_stats$Observation==obs & summary_stats$Treatment==treat,"grand_mean"]
        means<-summary_stats.pre[summary_stats.pre$Treatment==treat & summary_stats.pre$Observation==obs,"mean"]
        n<- length(means[!is.na(means)])
        summary_stats[summary_stats$Treatment==treat & summary_stats$Observation==obs,"grand_sd"]<-sqrt(sum((means-gm)^2,na.rm = T)/(n-1))
        summary_stats[summary_stats$Observation==obs & summary_stats$Treatment==treat,"n"]<-n
      }
    }
    
    sigtest<-apply_ttest(sigtest_results = sigtest,pval_threshold = pval_threshold,summary_stats = summary_stats,test_function = "t.test2")
    
    
  }else if(Experiment=="CONTROL_MEDS"){
    print("Nothing...")
  }
  
  else{
    print("Experiment must be Infection or Scratch")
  }
  
  
  statbars<-data.frame(xref=sigtest$xref,reference=sigtest$reference,
                       x=as.numeric(sigtest$xref)+.05,
                       xend=(as.numeric(sigtest$reference)+1)-.05)
  if(Experiment=="Scratch"){
    
    statbars[statbars$reference=="Cell Culture Medium","y"]<-rep(seq(100,91,-3),times=1)
    statbars[statbars$reference=="G. vaginalis","y"]<-rep(seq(91,by=-3,length.out = 3),times=1)
    statbars[statbars$reference=="L. iners","y"]<-rep(c(85,94),times=1)
    statbars[statbars$reference=="L. jensenii","y"]<-rep(94,times=1)
    statbars[statbars$reference=="Cell Culture Medium" & statbars$xref=="L. iners","y"]<-85
    statbars[statbars$reference=="G. vaginalis" & statbars$xref=="L. iners","y"]<-94
    
  }else if(Experiment=="Infection"){
    
    statbars[statbars$reference=="Cell Culture Medium","y"]<-rep(seq(100,97,-3),times=2)
    statbars[statbars$reference=="Fascaplysin","y"]<-rep(c(91,94),times=1)
    statbars[statbars$reference=="CAS 546102-60-7","y"]<-rep(90,times=2)
    statbars[statbars$xref=="CAS 546102-60-7" &statbars$reference=="Fascaplysin" ,"y"] <-97
    
  }else{
    print("Experiment must be Infection or Scratch")
  }
  
  
  statbars$yend<-statbars$y
  statbars2<-data.frame(x=c(statbars$x,statbars$xend),
                        y=rep(statbars$y,times=2),
                        yend=rep(statbars$y-1.5,times=1))
  statbars2$xend<-statbars2$x
  
  sigtest<-merge(statbars,sigtest,by = c("xref","reference"))
  
  sigtest$midpoints<-((sigtest$xend-sigtest$x)/2)+sigtest$x
  sigtest$y.sig<-sigtest$y+1
  sigtest$sig[is.na(sigtest$sig)]<-"N.S."
  
  return(list(sigtest=sigtest,statbars=statbars,statbars2=statbars2,summary_stats=summary_stats))
}

###########
## end setup_sigtest
###########

###########
## plot_replicates
###########


plot_replicates<-function(eset,BCS.selection=c(""),ExposureTime.selection=c(""),logt=F,rmlow=F,lowcnt=10){
  
  if(logt){
    cnts<-log(exprs(eset)+1,base = 2)
  }else{
    cnts<-exprs(eset)
  }
  if(rmlow){
    
    cnts<-cnts[rowSums(cnts>lowcnt)==ncol(cnts),]
  }
  
  pairs.panels(cnts[, pData(eset)$BCS %in% c(BCS.selection) & pData(eset)$ExposureTime %in% c(ExposureTime.selection)])
}


###########
## end plot_replicates
###########

###########
## subset_ExpressionSet
###########


```
# Temp
```{r tmp}

## SRL_meta and SRL_counts were prepared in separate script- 
load(file=paste0(R_script_input_directory,"SRL_counts_meta.RData"))

## Extract counts data from ExpressionSet
SRL_counts_table<-exprs(SRL_counts_meta)
SRL_meta_table<-pData(SRL_counts_meta)

## MOVE THIS TO PREP SCRIPT

taxa_data_dne<-is.na(SRL_meta_table$CST)
sum(taxa_data_dne)

#SRL_counts_meta<-subset_ExpressionSet(expSet = SRL_counts_meta,filterOut = SRL_meta_table[taxa_data_dne,"Pre_QC_ID"])

#SRL_counts_table<-exprs(SRL_counts_meta)
#SRL_meta_table<-pData(SRL_counts_meta)

##Write SRL Raw Counts to disk
write.csv(SRL_counts_table,file=paste0(thesis_tables_directory,TABLE_COUNTS_RAW),row.names=T,quote=F)
```

# Summary Statistics & Quality Control   

## Alignment Stats   
Summarize the SmallRNA seq alignments

```{r alignment_stats,eval=TRUE}

## Stats on # reads hg (% surviving), # reads miRNA (% survibing, hg)
SRL_meta_table_summary_table<-data.frame(matrix(0,nrow=3,ncol=3));names(SRL_meta_table_summary_table)<-c("min","median","max");row.names(SRL_meta_table_summary_table)<-c("trimmed","hg19","miRNome")

SRL_meta_table_summary_table["trimmed",]<-summarise(SRL_meta_table,min=min(reads_surviving),median=median(reads_surviving),max=max(reads_surviving))
SRL_meta_table_summary_table["hg19",]<-summarise(SRL_meta_table,min=min(hg19.mapped),median=median(hg19.mapped),max=max(hg19.mapped))
SRL_meta_table_summary_table["miRNome",]<-summarise(SRL_meta_table,min=min(number_reads_mirs),median=median(number_reads_mirs),max=max(number_reads_mirs))
print(SRL_meta_table_summary_table)

```

## Quality Control (QC) small RNA-seq count data

```{r QC,eval=TRUE}
print(paste("Number of samples in inital counts table:",ncol(SRL_counts_table)))
print(paste("Number of miRNAs in inital counts table:",nrow(SRL_counts_table)))

## Determine number of miRNAs without any counts across all samples 
missing<-rowSums(SRL_counts_table)==0 #rowSums(is.na(SRL_counts_table))==ncol(SRL_counts_table) | 
print(paste("Number of miRNAs without any counts across all",ncol(SRL_counts_table),"samples:",sum(missing)))
print(paste0("which leaves ",100*(1-sum(missing)/nrow(SRL_counts_table))," percent of all annotated miRNAs with at least 1 read"))

#print("...and the names of the miRNAs without any reads across all samples: ")
#names(missing)[missing]

plot_QC.preQC.batch<-plot_pca(SRL_counts_meta,"Log miRNA Raw Count Distance, PRE QC",color_by = "Batch",ploly = F,seed_val = seed_val)##FIGURE
plot(plot_QC.preQC.batch$pca.p)
plot_QC.preQC.batch$gPCA.result$p.val
plot_QC.preQC.sid<-plot_pca(SRL_counts_meta,"Log miRNA Raw Count Distance, PRE QC",color_by = "SID",ploly = F,seed_val = seed_val)##FIGURE
plot(plot_QC.preQC.sid$pca.p)
plot_QC.preQC.sid$gPCA.result$p.val
plot_pca(SRL_counts_meta,"Log miRNA Raw Count Distance, PRE QC",color_by = "BVGroup",seed_val = seed_val)$pca.p##FIGURE
plot_pca(SRL_counts_meta,"Log miRNA Raw Count Distance, PRE QC",color_by = "NugentC",seed_val = seed_val)$pca.p##FIGURE


## Get a *ROUGH* idea of miRNA coverage by dividing counts by # of annotated miRNAs in genome 
(number_mirs_genome<-nrow(SRL_counts_meta)) ## number of annotated miRNAs
#SRL_counts_table<-SRL_counts_table[!missing,] ## non missing counts only 
SRL_counts_meta<-SRL_counts_meta[!missing]
sum(missing)
## Number of miRNAs completley removed due to non-detection
nrow(SRL_counts_meta)

(plot_QC.RIN.percent.batch<-plot_RIN_meta(SRL_meta_table,"number_reads_mirs.percent_hgmapped",col_by="Batch",y_series_label = "Percentage of miRNA Reads of All hg19 Mapped Reads"))##FIGURE
(plot_QC.RIN.percent.sid<-plot_RIN_meta(SRL_meta_table,"number_reads_mirs.percent_hgmapped",col_by="SID",y_series_label = "Percentage of miRNA Reads of All hg19 Mapped Reads"))
(plot_QC.RIN.number.batch<-plot_RIN_meta(SRL_meta_table,"number_reads_mirs",col_by="Batch",y_series_label = "Number of miRNA Reads"))##FIGURE
(plot_QC.RIN.number.sid<-plot_RIN_meta(SRL_meta_table,"number_reads_mirs",col_by="SID",y_series_label = "Number of miRNA Reads"))

## Range of miRNA coverage based on simple reads/#annotated miRNAs
print(summary(colSums(SRL_counts_table))/number_mirs_genome)

## Avergae coverage
data.frame(Pre_QC_ID=SRL_meta_table$Pre_QC_ID,avg_coverage=SRL_meta_table$number_reads_mirs/number_mirs_genome)

hist(log(colSums(SRL_counts_table)/number_mirs_genome,10),main="Log10 Total miRNA Counts/# miRNAs")
abline(v = log(100,10))
.1*number_mirs_genome
## Samples with "low coverage", i.e., less than 1E4 reads/#miRs
low_coverage<-(colSums(SRL_counts_table))<=(125)*number_mirs_genome ##125
sum(low_coverage)
125 * 1869 
## An idea of the samples with "low coverage":
print(head(SRL_counts_table[,low_coverage]))

(removed_samples<-remove_poorQC_samples(removed_samples = removed_samples,sample_list =unique(colnames(SRL_counts_table[,low_coverage])),reason = "Low_Coverage"))

## Remove those samples with low coverage 
counts_meta.qc<-subset_ExpressionSet(expSet = SRL_counts_meta,filterOut = c(as.character(removed_samples$Pre_QC_ID)))

nrow(pData(counts_meta.qc))
ncol(exprs(counts_meta.qc))

ncol(pData(counts_meta.qc))
nrow(exprs(counts_meta.qc))

# ## Re-plot after QC
plot_QC.postQC.batch<-plot_pca(counts_meta.qc,"Log miRNA Raw Count Distance, POST QC",color_by ="Batch",ploly = F,seed_val = seed_val)##FIGURE
plot_QC.postQC.batch$pca.p

plot_QC.postQC.sid<-plot_pca(counts_meta.qc,"Log miRNA Raw Count Distance, POST QC",color_by = "SID",ploly = F,seed_val = seed_val)##FIGURE
plot_QC.postQC.sid$pca.p

plot_pca(counts_meta.qc,"Log miRNA Raw Count Distance, POST QC",color_by = "BVGroup",ploly = F,seed_val = seed_val)$pca.p##FIGURE
plot_pca(counts_meta.qc,"Log miRNA Raw Count Distance, POST QC",color_by = "NugentC",ploly = F,seed_val = seed_val)$pca.p##FIGURE

plot_RIN_meta(pData(counts_meta.qc),"hg.mapped.percent",col_by = "Batch")##FIGURE
plot_RIN_meta(pData(counts_meta.qc),"number_reads_mirs.percent_hgmapped",col_by = "Batch")##FIGURE
plot_RIN_meta(pData(counts_meta.qc),"number_reads_mirs.percent_surviving",col_by = "Batch")##FIGURE
dev.off()
median_RIN<-median(pData(counts_meta.qc)$RIN,na.rm = T)

print(paste0("The median RINe score was ",median_RIN,", with ",round(100*sum(pData((counts_meta.qc))$RIN<7,na.rm = T)/length(pData((counts_meta.qc))$RIN),digits = 1),"% of the samples having a RINe greater than 7."))

number_miRNAs<-mapping_stats(pData(counts_meta.qc)$number_reads_mirs)
percent_miRNAs.hg<-round(mapping_stats(pData(counts_meta.qc)$number_reads_mirs.percent_hgmapped),1)
percent_miRNAs.oftrimmed<-round(mapping_stats(pData(counts_meta.qc)$number_reads_mirs.percent_surviving),1)

print(paste0("The median (minimum/maximum) percentage of post-QC miRNA reads relative to all hg19 mapped and total reads was ",percent_miRNAs.hg$median,"% (",percent_miRNAs.hg$min,"%/",percent_miRNAs.hg$max,"%) and ",percent_miRNAs.oftrimmed$median,"% (",percent_miRNAs.oftrimmed$min,"%/",percent_miRNAs.oftrimmed$max,"%), respectively. However, the  median number of post-QC hg19 mapped miRNA reads was ",number_miRNAs$median,", with a minimum ",number_miRNAs$min," and maximum ",number_miRNAs$max,". Thus, despite low relative miRNA read counts, the estimated coverage ranged from ",round(number_miRNAs$min/number_mirs_genome,digits = 0),"X-",round(number_miRNAs$max/number_mirs_genome,digits = 0),"X across the entire miRnome (",number_mirs_genome," annotated miRNAs)."))

sum(pData((counts_meta.qc))$number_reads_mirs<10^6,na.rm = T)/length(pData((counts_meta.qc))$number_reads_mirs)

##FIXME

# cairo_ps(file = paste0(thesis_figures_directory,FIGURE_QC_PCA.PREQC.BYBATCH),width = 8,height = 5.5)
# plot(plot_QC.preQC.batch$pca.p)
# dev.off()
# cairo_ps(file = paste0(thesis_figures_directory,FIGURE_QC_PCA.PREQC.BYSUBJ),width = 8,height = 5.5)
# plot(plot_QC.preQC.sid$pca.p)
# dev.off()
# cairo_ps(file = paste0(thesis_figures_directory,FIGURE_QC_PCA.RMLOW.BYBATCH),width = 8,height = 5.5)
# plot(plot_QC.postQC.batch$pca.p)
# dev.off()
# cairo_ps(file = paste0(thesis_figures_directory,FIGURE_QC_PCA.RMLOW.BYSUBJ),width = 8,height = 5.5)
# plot(plot_QC.postQC.sid$pca.p)
# dev.off()
# cairo_ps(file = paste0(thesis_figures_directory,FIGURE_QC_RIN_v_READS.PROP),width = 8,height = 5.5)
# plot(plot_QC.RIN.percent.batch)
# dev.off()
# cairo_ps(file = paste0(thesis_figures_directory,FIGURE_QC_RIN_v_READS.ABS),width = 8,height = 5.5)
# plot(plot_QC.RIN.number.batch)
# dev.off()

```
# Discovery of miRNAs Associated w/ Lactobacillus spp. vs CST-IV/BV  

## Normalize & log2 transform small RNA-Seq counts for use in models   

```{r normalize,eval=TRUE,warning=FALSE}
## Calcualte size factors for normalization using calcNormFactors function
hist(log(exprs(counts_meta.qc),base = 2))
expSet.sizeDisp<-calcNormFactors(method = "TMM",DGEList(counts = exprs(counts_meta.qc)),logratioTrim = 0.10,sumTrim = 0.05) 

## Normalize counts using size dispersions and norm factors, store as Expression Set
expSet.normalized<-ExpressionSet(assayData = as.matrix(1E6*expSet.sizeDisp$counts*expSet.sizeDisp$samples$norm.factors/expSet.sizeDisp$samples$lib.size),phenoData = AnnotatedDataFrame(pData(counts_meta.qc)))

## Store log-transformed QC *NON NORMALIZED* count table for PCA. This is not used for anything but PCA. 
expSet_log<-ExpressionSet(assayData = log(exprs(counts_meta.qc),base=2),phenoData = AnnotatedDataFrame(pData(counts_meta.qc)))

## Store log-transformed QC normalized count table. This is the small RNAseq count table used in all downstream analysis
expSet_log.normalized<-ExpressionSet(assayData = log(exprs(expSet.normalized)+1,base=2),
                                     phenoData = AnnotatedDataFrame(pData(expSet.normalized))) ## add pseuo counts... so that when take log(), all 0 reads become 0 normalized reads log(raw + 1)=log(1)=0. LIttle effect on other counts. This makes it esier to drop raw 0 counts later. Also store count table annotation from original Expression Set


# ## PCA plots after normalization, colored by batch, SID and "BV group"
plot_QC.normalized.batch<-plot_pca(expSet.normalized,color_by = "Batch",margins = unit(c(0,-200,0,-200),units = "points"),seed_val = seed_val)
plot_QC.normalized.batch$pca.p
plot_QC.normalized.sid<-plot_pca(expSet_log.normalized,color_by = "SID",ploly = F,logt = F,margins = unit(c(0,30,0,5),units = "pt"),seed_val = seed_val)
plot_QC.normalized.sid$pca.p
plot_pca(expSet_log.normalized,color_by = "BVGroup",logt = F)$pca.p
plot_pca(expSet_log.normalized,color_by = "NugentC",logt = F)$pca.p
# 
# ## Print for text
print(paste0("PCA plots before and after sample removal and after normalization do not support batch effects (guided PCA p-values ",plot_QC.preQC.batch$gPCA.result$p.val,", ",plot_QC.postQC.batch$gPCA.result$p.val,", and ",plot_QC.normalized.batch$gPCA.result$p.val,", respectively), or subject-specific effects (guided PCA p-values ",plot_QC.preQC.sid$gPCA.result$p.val,", ",plot_QC.postQC.sid$gPCA.result$p.val,", and ",plot_QC.normalized.sid$gPCA.result$p.val,", respectively, Figure S1)."))

## Store counts data as "features"
features<-exprs(expSet_log.normalized)
head(features)[,1:3]

## Replace NA values with half the minimum normalized count values so that modelling works with algorithms
(min_normalized_count<-2^min(features,na.rm = T))
features[is.na(features)]<-log(min_normalized_count/2,base = 2)

## Remove low count miRNAs
hist(features,main="Histogram of Normlized Read Counts",xlab="Log (Normlized read count)")
abline(v = -2,col='red')
nrow(features)
keep_features<-rowSums(features>log(min_normalized_count/2,base = 2))>=(.5*ncol(features))
sum(keep_features)
features<-features[keep_features,] ## At least half of normalized miRNA counts should be >1. 
hist(features,main="Histogram of Normlized Read Counts",xlab="Log (Normlized read count)")
abline(v = 1,col='red')

##FIXME

# ## Write PCA plots to file
cairo_ps(file=paste0(thesis_figures_directory,FIGURE_QC_PCA.NORMAL.BYBATCH),width=8,height=5.5)
plot_QC.normalized.batch$pca.p
dev.off()
cairo_ps(file=paste0(thesis_figures_directory,FIGURE_QC_PCA.NORMAL.BYSUBJ),width=8,height=5.5)
plot_QC.normalized.sid$pca.p
dev.off()

```

## Prepare & Analyze the Nugent input tables (predictors) to RF model   

```{r prepare_model_input,eval=T}
load(file=paste0(R_script_input_directory,"CV_16S_AMSEL.Rdata"))
load(file=paste0(R_script_input_directory,"SRL_16S.Rdata"))
## FIXME make sure the miRNA data intersects with the 16S data!

SRL_16S<-SRL_16S[!row.names(SRL_16S) %in% as.character(removed_samples$Pre_QC_ID),] 

LACTO_PROP<-rowSums(dplyr::select(SRL_16S,c(Lactobacillus_crispatus,Lactobacillus_jensenii,Lactobacillus_iners,Lactobacillus_gasseri)))

## Join the existing metdadata and predicted Amsel diagnosis
joined_meta_amselPred<-join(pData(counts_meta.qc),data.frame(Pre_QC_ID=names(LACTO_PROP),LACTO_PROP=LACTO_PROP),by="Pre_QC_ID")
row.names(joined_meta_amselPred)<-joined_meta_amselPred$Pre_QC_ID


##Store the joined data as features_metadata
features_metadata<-ExpressionSet(features,phenoData = AnnotatedDataFrame(joined_meta_amselPred))

## Summarize postQC counts data/metadata
summary_sid.group<-ddply(pData(counts_meta.qc),c("SID","BVGroup"),summarise,N=length(SID))
(summary_group.sid<-ddply(pData(counts_meta.qc),c("BVGroup","SID"),summarise,N=length(SID)))
(summary_group<-ddply(pData(counts_meta.qc),c("BVGroup"),summarise,N=length(SID)))
(summary_group.uniqSID<-ddply(unique(dplyr::select(pData(counts_meta.qc),c(SID,BVGroup))),c("BVGroup"),summarise,N=length(SID)))
(summary_group.race<-ddply(unique(dplyr::select(pData(counts_meta.qc),c(BVGroup,Race,SID))),c("BVGroup","Race"),summarise,N=length(SID)))

print(paste0("A total of ",sum(summary_sid.group$N)," samples, representing ",sum(summary_group.uniqSID$N)," unique subjects from one of 3 longitudinal groups were used in the final analysis"))

## Subset metadata for easier handling downstream
selected_metadata<-dplyr::select(pData(features_metadata),c(SID,NUGENT_SCORE,LACTO_PROP,Pre_QC_ID,CST)) 
#selected_metadata<-dplyr::select(pData(features_metadata),c(SID,VAG_INT,FING_PEN,contains("SEX"),symptoms,symptoms_nonBV,BirthControl,NUGENT_SCORE,LACTO_PROP,Pre_QC_ID,MENSTRUATION_NORMALIZED_PHASED,CST)) ## CSTs are *NOT* used in model input, but are instead passed along to top mir expression plot/table downstream. Remove CST from running in model.  

##Store only counts expressed 'above 0' to reduce noise in RF models
min(exprs(features_metadata))
expressed_above_0<-exprs(features_metadata)[rowSums(exprs(features_metadata)!=0)>=(1*ncol(exprs(features_metadata))),] 

## Combine non zero log transformed normalized counts with selected metadata as input into RF models (predictors + responses)
model_input<-data.frame(t(expressed_above_0),selected_metadata)
row.names(model_input)<-model_input$Pre_QC_ID
model_input<-dplyr::select(model_input,-Pre_QC_ID)
model_input$LACTO_PROP<-as.numeric(model_input$LACTO_PROP)
names(model_input)<-gsub(gsub(gsub(names(model_input),pattern = "mir.",replacement = "miR-"),pattern = "hsa.",replacement = ""),replacement ="-",pattern = "\\.")


model_input_Lactobacillus<-dplyr::select(model_input[!is.na(model_input$LACTO_PROP),],-c(NUGENT_SCORE,CST))

samples_without_16S<-setdiff(row.names(model_input),row.names(model_input_Lactobacillus))

model_input_Lactobacillus[is.na(model_input_Lactobacillus)]<-0 ## RF can not have missing values. Replacing with 0's isn't technically valid, but should have a minimal effect on outcome

##ADD QC STAGE TO SRL_meta_table, then write to disk
SRL_meta_table<-left_join(SRL_meta_table,removed_samples,by="Pre_QC_ID")
print(paste0("There were 5 samples removed due to insufficient library material or failure to sequence, and ",sum(!is.na(SRL_meta_table$QC_removal_stage) & SRL_meta_table$QC_removal_stage %in% c("Low_Coverage","Visual Outlier"))," samples were removed due to low total miRNA reads or outliers."))

write.csv(SRL_meta_table,file=paste0(thesis_tables_directory,TABLE_SRL_METADATA),row.names=F,quote=F)

## Summarize # samples surviving QC and having 16S data
SRL_seq_summary<-ddply(SRL_meta_table,c("BVGroup","SID"),summarise,PreQC=length(QC_removal_stage),PostQC=length(QC_removal_stage)-sum(as.numeric(QC_removal_stage),na.rm = T))

## Count # samples with 16S data
samples_with_16S.nsid<-data.frame(table(dplyr::filter(SRL_meta_table,!Pre_QC_ID %in% samples_without_16S & is.na(SRL_meta_table$QC_removal_stage))$SID))
names(samples_with_16S.nsid)<-c("SID","With Lact. RA")
SRL_seq_summary<-merge(SRL_seq_summary,samples_with_16S.nsid,all.x = T)

## Write to table
SRL_seq_summary<-SRL_seq_summary[order(SRL_seq_summary$BVGroup),]
write.csv(SRL_seq_summary,file=paste0(thesis_tables_directory,TABLE_SEQSUMMARY),row.names=F,quote=F)

```
## Run Random Forest Models   
Use RF to discover miRNAs associated w/ Lactobacillus spp. dominated communities vs CST-IV (BV-associated communities)   
* Amsel- Random Forest (using predicted Amsel diagnoses as response variable)  
* Nugent- Random Forest (using Nugent score as response variable)   

```{r run_model,eval=T}
## //////////
### Random Forest
## //////////
## See run_randomForest function above for paramaters. Note that you can load a previously-built model to save time.  Note that if loaded from previous, most of the input is ignored. 

##FIXME change all Nugent to Lacto (or just RF) and remove Amsel. 
model_input_Lactobacillus

setdiff(c(Lacto_RF$training_ids,Lacto_RF$testing_ids),row.names(model_input_Lactobacillus))

Lacto_RF<-run_randomForest(predictors_response_table = model_input_Lactobacillus,
                           response_variable_name = "LACTO_PROP",
                           subj_spec =T,
                           nfold = nfolds,
                           nreps = npermutes,
                           permute = T,
                           save_model = T,
                           file_n = "Lacto_RF_new2",
                           load_prev_model = F,
                           verbose = F,
                           training_prop = training_prop,
                           R_script_output_directory=R_script_output_directory) 

## Store top miRNAs
(Lacto_RF$top_features$top_features.all<-row.names(Lacto_RF$rfp$importance_w_pval[Lacto_RF$rfp$importance_w_pval[,"%IncMSE.pval"]<=pval_threshold & Lacto_RF$rfp$importance_w_pval[,"IncNodePurity.pval"]<=pval_threshold ,]))

##Calculate accuracy/mean abs error
accuracy_table.Nugent<-Lacto_RF$accuracy_table
(mean_absolute_error.Nugent<-sum(abs(accuracy_table.Nugent$predicted- accuracy_table.Nugent$actual))/nrow(accuracy_table.Nugent))
hist(abs(accuracy_table.Nugent$predicted- accuracy_table.Nugent$actual))
plot(accuracy_table.Nugent$predicted, accuracy_table.Nugent$actual)
print(paste0("The Nugent-RF model correctly predicted the Nugent score within ",round(mean_absolute_error.Nugent,digits = 2)," values on average. There were ",length(Lacto_RF$top_features$top_features.all)," significant miRNAs using Nugent-RF."))

## //////////
### Output Model Results
## //////////

### Tables
## Write model input table to file, keeping only variables that were used 
model_input.out<-model_input[,names(model_input) %in% unique(c(names(model_input_Lactobacillus)))]

predictor_names<-names(dplyr::select(model_input.out,-c(LACTO_PROP,SID)))
predictor_names.miRs<-sum(grepl(predictor_names,pattern = "miR|let"))
print(paste0("The Lacto-RF models used ",length(predictor_names)," predictors including ",predictor_names.miRs," non-zero log2 transformed miRNA read counts and ",length(predictor_names)-predictor_names.miRs," metadata variables as inputs to rank feature importance (",TABLE_MODEL_INPUT,")."))

## Record which variables were used in training and testing
model_input.out[row.names(model_input.out) %in% Lacto_RF$training_ids,"Lacto_RF_trainingTestingSet"]<-"training"
model_input.out[row.names(model_input.out) %in% Lacto_RF$testing_ids,"Lacto_RF_trainingTestingSet"]<-"testing"

write.csv(model_input.out,paste0(thesis_tables_directory,TABLE_MODEL_INPUT),row.names=T,quote=F)

## Output RF model results
TABLE_RF_SUMMARY_output<-data.frame(miRNA=row.names(Lacto_RF$rfp$importance_w_pval),Lacto_RF$rfp$importance_w_pval)

write.csv(TABLE_RF_SUMMARY_output,file=paste0(thesis_tables_directory,TABLE_RF_SUMMARY),row.names=F,quote=F)

##Store the top miRNAs for later use
top_mirs_table<-data.frame(miRNA=Lacto_RF$top_features$top_features.all)

## Text
print(paste0("A total of ",length(top_mirs_table$miRNA)," miRNAs were identified uing Lactobacillus RA.: ", str_c(top_mirs_table$miRNA,collapse=", "),"."))


### Figures
Lacto_RF.plot<-plot_importance(dplyr::select(data.frame(Lacto_RF$rfp$importance_w_pval)[row.names(Lacto_RF$rfp$importance_w_pval) %in% top_mirs_table$miRNA,],contains("MSE")),ntopfeats =  11,rankBy = "IncMSE",nfeats = 11,model_name = "Lacto",size_font = 20)

plot(Lacto_RF.plot)

cairo_ps(file=paste0(thesis_figures_directory,FIGURE_RF_IMPORTANCE_AMSEL), width=11, height=8.5)
plot(Lacto_RF.plot)
dev.off()


topmir_plot_data<-model_input[!is.na(model_input$NUGENT_SCORE),names(model_input) %in% c("NUGENT_SCORE","LACTO_PROP","CST",as.character(top_mirs_table$miRNA))]
topmir_plot_data$CST<-as.character(topmir_plot_data$CST)

topmir_plot_data<-melt(topmir_plot_data,id.vars = c("NUGENT_SCORE","LACTO_PROP","CST"))

topmir_plot_data$variable<-gsub(topmir_plot_data$variable,pattern = "hsa.mir.",replacement = "miR-")

## Calcualte linear model fit for miRNA expression data vs Nugent score to rank miR expression plots in figure

fitness<-apply(model_input[,names(model_input) %in% unique(topmir_plot_data$variable)],MARGIN = 2,function(s) summary(lm(model_input$NUGENT_SCORE~s)))

fitness.r2<-data.frame(adjr2=unlist(lapply(fitness,function(x) round(x$adj.r.squared,digits = 3))))
fitness.r2$variable<-row.names(fitness.r2)

print(fitness.r2[order(abs(fitness.r2$adjr2),decreasing = T),])

topmir_plot_data<-join(topmir_plot_data,fitness.r2)

topmir_plot_data$variable<-factor(x=topmir_plot_data$variable,levels = unique(topmir_plot_data[order(abs(topmir_plot_data$adjr2),decreasing = T),c("variable")]),ordered = T)

## Liner correlation coeff for miRNA expression data:
paste(apply(unique(dplyr::select(topmir_plot_data,c(variable,adjr2))),MARGIN = 1,FUN = function(x) paste(sep =":" ,x[1],x[2])),sep=",")

## Collapse CSTs to a single CST super type
topmir_plot_data[topmir_plot_data$CST %in% c("III-A","III-B"),"CST"]<-"III"
topmir_plot_data[topmir_plot_data$CST %in% c("I-A","I-B"),"CST"]<-"I"

##FIXME make sure NA samples weren't use in RF-- samples without a lacto prop. 

## Plot the miRNA expression figure
PLOT_EXPRESSION<-
  ggplot(topmir_plot_data,aes(x=100*LACTO_PROP,y=value))+
  stat_smooth(method = "lm",col='#FF5733',se=F)+ #
  geom_point(aes(col=as.factor(CST)),size=3,show.legend = T)+
  facet_wrap(~variable,scales = "free_y",nrow=2)+
  ylab("Normalized Expression")+xlab("Lactobacillus spp. relative abundance (%)")+
  journalFormat+
  theme(text = element_text(size=20),axis.text.x = element_text(size=15))+
  scale_x_continuous(breaks=seq(from = 0,to=100,by=20))+
  scale_color_manual(values = cst.colors,guide = guide_legend(title = "CST"))

cairo_ps(paste0(thesis_figures_directory,FIGURE_TOPMIRS),width = 16,height = 8) #
plot(PLOT_EXPRESSION)
dev.off()


library(Boruta)

postscript(paste0(R_script_output_directory,"Boruta_optimization.ps"))
Lacto_RF.Boruta<-Boruta(x=dplyr::select(model_input_Lactobacillus,-c(SID,LACTO_PROP)),y=model_input_Lactobacillus$LACTO_PROP)
plot(Lacto_RF.Boruta,main="Nugent-RF Boruta unimportant feature removal")
Lacto_RF.Boruta$finalDecision[Lacto_RF.Boruta$finalDecision=="Confirmed"]


dev.off()
```


## Map miRNAs to Gene Targets and GO Process (miR-GO-Target) (Table 2)

Using experimentally validated miRNA targets, map to direct GO processes to discover common & unique functions of top miRNAs. 
1.   Output gene target list for each top miRNA.    
2.   Count the frequency of each target for each GO process & miRNA. Plot. 
```{r miRGoTarget,eval=T}

### miRNA <-> Taget (gene) <-> Gene Ontology Process (direct)
### miRNA <- - - - - - - -- >  Gene Ontology Process (direct)

## Read in miRTarBase Experimentally Validated Target List
mir_targets<-read.table(paste0(R_script_input_directory,"miRTarBase_SE_WR_homosapiens.txt"),header = T,sep="\t",stringsAsFactors = F) %>% filter(Species..miRNA.=="Homo sapiens")
nrow(mir_targets)

## Format targets list,pattern = "hsa.",replacement = "")
mir_targets$miRNA.format<-gsub(gsub(gsub(mir_targets$miRNA,pattern = "mir.",replacement = "miR-"),replacement ="-",pattern = "\\."),pattern = "-.p",replacement = "-")

## Map miRNA names to target List
for(mir in as.character(top_mirs_table$miRNA)){
  
  target_list<-mir_targets[grep(mir_targets$miRNA.format,pattern = paste0("hsa-",mir,"-"),ignore.case = T,perl = T),c("miRNA","Target.Gene","References..PMID.")]
  #if(!nrow(target_list)==0){target_list$source<-"canonical"}
  if(!nrow(target_list)==0 ){
    top_mirs_table[top_mirs_table$miRNA==mir,"targets"]<-str_c(sort(unique(target_list$Target.Gene)),collapse = ",")
    top_mirs_table[top_mirs_table$miRNA==mir,"sanityCheck"]<-str_c(sort(unique(target_list$miRNA)),collapse = ",")
    top_mirs_table[top_mirs_table$miRNA==mir,"PMID"]<-str_c(sort(unique(target_list$References..PMID.)),collapse = ",")
  }
}

write.table(top_mirs_table,file=paste0(thesis_tables_directory,TABLE_MIR_TARGETS),sep="\t",quote = F,row.names = F) ## Needs to be tab delimited due to "," separating gene targets. 

### Read in miR-Go-Gene mapping table mapping targets/miRNAs to GO processes. Gene targets mapped to miRNAs using a Python script. 
mir_go<-read.csv(paste0(R_script_input_directory,"miR_GODirect_Gene_Map.csv"),stringsAsFactors = F) 


## Filter just the miR-GO map to 'top' miRNAs discovered above belonging to both RF model results 
mir_go<-mir_go[mir_go$mir %in% as.character(top_mirs_table$miRNA),]
ddply(unique(dplyr::select(mir_go,c(mir,gene))),c("mir"),summarise,n=length(gene))

## Read in miR-target counts, created in outside script. 
mir_targetCnt<-read.csv(paste0(R_script_input_directory,"miR_Target_Counts.csv"),stringsAsFactors = F)
mir_targetCnt[mir_targetCnt$mir %in% top_mirs_table$miRNA,]
## Determine the number of genes associated with each miRNA- need to determine how many GO processes to plot in figure
mir_go.counts<-data.frame(table(dplyr::select(mir_go,c(mir,GO))))
mir_go.miR193b<-mir_go[mir_go$mir=="miR-193b",]

mir_go.counts<-merge(mir_go.counts,mir_targetCnt)
mir_go.counts$Freq_Targets<-mir_go.counts$Freq/mir_go.counts$number_targets

hist(mir_go.counts$Freq_Targets[!mir_go.counts$Freq_Targets==0],main="Historgam of Non-zero miRNA Target Frequency",xlab="Proportion of Targets")

## Somewhere above 0.2 might capture the interesting/common GO processes across miRNAS. 

## Make miR-GO counts into wide format (i.e., each GO process is a column, miRNAs are rows, values are proportion of targets in that GO/miRNA combination)
mir_go.wide<-tidyr::spread(dplyr::select(mir_go.counts,-c(Freq,number_targets)),key = GO,value = Freq_Targets)

row.names(mir_go.wide)<-mir_go.wide$mir
mir_go.wide<-dplyr::select(mir_go.wide,-mir)

## Get a better idea of how to plot common GO processes by determining cutoffs for target proportions
## Balanace between proportion of targets across all miRNAs and total # across targets

## create dummy dataframe- i and j hold the tuning parameters
df1<-data.frame(i=rep(seq(0,max(mir_go.wide),max(mir_go.wide)/10),each=10),j=rep(seq(0,1,.1),times=10))

## Calcualte number of columns (nc) for i and j cutoffs
for(a in 1:nrow(df1)){
  i<-df1[a,"i"]
  j<-df1[a,"j"]
  ## number of columns w/ propotion >i and of those, number greater than j of total miRNAs
  nc<-ncol(mir_go.wide[,colSums(mir_go.wide>i)>=j*nrow(mir_go.wide)])
  if(is.null(nc)){nc<-0}
  df1[a,"ncol"]<-nc
}
(optimal_cutoffs<-df1[df1$ncol==max(df1[df1$ncol<=25 & df1$ncol>=5,"ncol"]),]) ## What are the optimal cutoffs for somewhere between 5 and 25 GO processes on the plot?

## Take the max i/j pair as the cutoffs:
i<-.1125#0.1904762 # 0.2285714 #0
j<-0.6#0.3 #0.8
## Note that 19 in the "nc" column is the number of GO processes that will be plotted

## Therefore, the proportion of genes should be >0 across at least 0.8*6 =4.8 miRNAs. 
hist(as.matrix(mir_go.wide))

## Apply optimal cutoffs to miR-GO (wide) data
mir_go.wide.filtered<-mir_go.wide[,colSums(mir_go.wide>i)>=j*nrow(mir_go.wide)]

mir_go.wide.mir193b<-mir_go.wide[row.names(mir_go.wide)=="miR-193b",]

## The names of the GO processes to be used in plot:
names(mir_go.wide.filtered[,order(colSums(mir_go.wide.filtered>0),decreasing = T)])

mir_go.wide.filtered$mir<-row.names(mir_go.wide.filtered)
mir_go.wide.filtered.m<-melt(mir_go.wide.filtered,id.vars = c("mir"))
mir_go.wide.filtered.m$GO_only<-gsub(mir_go.wide.filtered.m$variable,pattern = "~.*",replacement = "") ## Store the GO number only in case needed

## For this plot, summarize the mean gene target proprotion and # of miRNAs for each GO process. Use this to rank GO processes in plot
(mir_go.wide.filtered.summary<-ddply(mir_go.wide.filtered.m,c("variable"),summarise,s=sum(value),n=sum(value>0)))

## TO make plot simplier to read, group GO terms in logical super groupings. This is done using a map between GO group and GO term, created outside of script
group_go_map<-read.csv(file=paste0(R_script_input_directory,"GO_Process_Grouping_Map.csv"))

## Join the map with the GO process summary table. Note that the map may contain more GO processes than the summary, due to slight changes in final miRNA list. 
group_go_map<-merge(group_go_map,mir_go.wide.filtered.summary)

## How many GO terms in each super group
(group_freq<-data.frame(table(group_go_map$Group))) 
names(group_freq)<-c("Group","n_group")
# MErge the super groups and data found above
group_go_map<-merge(group_go_map,group_freq)

## Order by grouping, then by mean target proportion within each group
group_go_map<-group_go_map[order(group_go_map$n_group, group_go_map$s,decreasing = T),]

#FOrmat the GO term for plot
mir_go.wide.filtered.m$GO_format<-factor(mir_go.wide.filtered.m$variable,levels =group_go_map$variable ,ordered = T,labels = group_go_map$Go_format)

## Determine order of miRs on plot according to correlation with Lactoabacillus spp prop
miR_order<-unique(dplyr::select(topmir_plot_data,c(variable,adjr2)))
miR_order<-miR_order[order(miR_order$adjr2,decreasing = T),]
mir_go.wide.filtered.m$mir<-factor(mir_go.wide.filtered.m$mir,levels = miR_order$variable,ordered = T)

PLOT_MIR_TARGET_GO_PROP<-ggplot(mir_go.wide.filtered.m,aes(x=GO_format,y=100*value,fill=mir))+geom_bar(width=.7,position = "dodge",col="grey",stat="identity")+
  journalFormat+
  theme(axis.text.x = element_text(angle=45,hjust = 1,size=10),text=element_text(size=12),plot.margin = unit(c(10,25,10,170),units = "pt"))+
  scale_fill_brewer(type = "qual",palette = "Set1")+
  xlab("GO Term")+
  ylab("Percentage miRNA Targets")

cairo_ps(paste0(thesis_figures_directory,FIGURE_MIR_TARGETS_GO),width = 11,height = 8.5)
plot(PLOT_MIR_TARGET_GO_PROP)
dev.off()
plot(PLOT_MIR_TARGET_GO_PROP)

## NOTE: the # of miR targets table is added later. See above for table. 

## Output miR-Target-GO table for miR-193b only
mir_go.wide.filtered.m.193b<-filter(mir_go.wide.filtered.m,mir=="miR-193b")

write.table(dplyr::select(mir_go.wide.filtered.m.193b,c(mir,GO_format,value)),file = paste0(thesis_tables_directory,TABLE_TOPMIRS),sep="\t",quote=F,row.names = F)

## Make a GO process bar chart for miR-193b only (for thesis presentation, not used in thesis document)
mir193b_GOcount_groups<-data.frame(table(dplyr::select(unique(dplyr::select(merge(mir_go.miR193b,group_go_map,by.x="GO",by.y="variable"),c(Group,gene))),Group)))
mir193b_GOcount_groups$Var1<-factor(mir193b_GOcount_groups$Var1,levels=mir193b_GOcount_groups[order(mir193b_GOcount_groups$Freq,decreasing = T),"Var1"],ordered = T)

mir193b_GOcount_groups.plot<-ggplot(mir193b_GOcount_groups,aes(x=Var1,y=Freq))+geom_bar(width=.7,position = "dodge",col="grey",stat="identity")+
  journalFormat+
  theme(axis.text.x = element_text(angle=45,hjust = 1),text=element_text(size=25),plot.margin = unit(c(10,25,10,170),units = "pt"))+
  scale_fill_brewer(type = "qual",palette = "Set1")+
  scale_y_continuous(breaks=c(2,4,6,8))+
  xlab("GO process")+
  ylab("Number miR-193b targets")

cairo_ps(paste0(R_script_output_directory,"miR193b_GOProcess_presentation.eps"),width = 11,height = 8.5)
plot(mir193b_GOcount_groups.plot)
dev.off()

```

# Function of miR-193b and in vitro Experimentation, Implications  

Perform further experimentation on miR-193b, one of the top miRNAs found in the discovery phase above.

## Validate SmallRNASeq Results using qPCR
::select 5 swabs each from NBV and PBV subjects, then measure relative hsa-mir-193b expression   

```{r qPCR_Validation_SRL,eval=T}
## Loads qPCR Data from 5 NBV and PBV subjects to Validate miRNA-Seq
load(paste0(R_script_input_directory,"miRNA_qPCR_Validation.RData"))
miR_qPCR_results.deltaCt<-miR_qPCR_results.deltaCt[!is.nan(miR_qPCR_results.deltaCt$deltaCt),] ## remove NA from data

## Create a mixed-effects linear model as a function of BVGroup of dCt values, using subject ID as the random effect
seq_validation.lme <- lme(deltaCt ~ BVGroup,
                          random = ~ 1|SID,
                          data = miR_qPCR_results.deltaCt)

## Coefficient of PBV relative to NBV in model + p value
effect_PBV<-seq_validation.lme$coefficients$fixed["BVGroupPBV"]
dct.summary<-ddply(miR_qPCR_results.deltaCt,c("BVGroup"),summarise,mean=mean(deltaCt))
2^(dct.summary$mean[2]-dct.summary$mean[1])
summary(seq_validation.lme)

```

## miR-193b qPCR Time-Course in VK2 monolayer after BCS Exposure
qPCR is measured with QIAGEN miScript II kit, BCSs are:   
*  L. crispatus  
*  L. jensenii  
*  L. iners  
*  G. vaginalis   
*  Media control    

20% BCSs exposed to VK2 monolayer epithelial cells for:   
* 0.5 hours   
* 1 hour   
* 4 hours   
* 8 hours   
* 11 hours   
* 13 hours  
* 16 hours   
* 19 hours   
* 22 hours   

Primers:   
* hsa-miR-193b-3p   
* hsa-miR-RNU6 (normalization control)   
```{r qPCR_time_course,eval=T}

## Breadcrumb: 
# qPCR_time_course.Rdata and qpcr_sigtest.Rdata from PreparePipeline/Prepare_qPCR.Rmd, previously called PreparePipeline/BCM_Experiments.Rmd (archives in external HD or Github may be named this)

# Calculate mean & s.d. Ct values for each BCS/time point. 
load(file=paste0(R_script_input_directory,"qPCR_time_course.Rdata")) ## Ct values
load(file=paste0(R_script_input_directory,"qpcr_sigtest.Rdata")) ## Container for holding sig testing comparisons/results. Generated so that different studies are matched with their respective controls. 

qPCR_time_course.summary<-ddply(qPCR_time_course,c("Study","ExposureTime","BCS"),summarise,m_193b=mean(miR193b_Ct,na.rm = T),sd_193b=sd(miR193b_Ct,na.rm = T),
                                m_R6=mean(RNU6_Ct,na.rm = T),sd_R6=sd(RNU6_Ct,na.rm = T))
##qPCR_time_course.summary m, sd of Cts 
qPCR_time_course$ExposureTime.n<-as.numeric(gsub(qPCR_time_course$ExposureTime,pattern = "hr",replacement = ""))

## Calcualte delta Ct using RNU6 as endogenous control. Calcualte standard deviation using propagation of error
qPCR_time_course.summary$deltaCt.RNU6<-qPCR_time_course.summary$m_193b-qPCR_time_course.summary$m_R6
qPCR_time_course.summary$deltaCt.RNU6.sd<-sqrt(qPCR_time_course.summary$sd_193b^2+qPCR_time_course.summary$sd_R6^2)


rm_rows<-qPCR_time_course.summary[is.na(qPCR_time_course.summary$sd_193b),c("ExposureTime","BCS")]
qpcr_sigtest<-qpcr_sigtest[!(qpcr_sigtest$comparison %in% rm_rows$BCS & qpcr_sigtest$ExposureTime %in% rm_rows$ExposureTime),]

## calcualte signifigance using delta Ct mean + s.d., which uses a modified t.test2 function. 
## Loop through the comparisons and test pair-wise
for(sigtest in 1:nrow(qpcr_sigtest)){
  timei<-as.character(qpcr_sigtest[sigtest,"ExposureTime"]) # TIme point
  x1<-as.character(qpcr_sigtest[sigtest,"comparison"]) ## Comparison sample
  y1<-as.character(qpcr_sigtest[sigtest,"reference"]) ## Reference sample
  
  ## When comparing samples, make sure to use the correct study associated with BCS
  study.x<-as.character(qpcr_sigtest[sigtest,"Study.comp"])
  study.y<-as.character(qpcr_sigtest[sigtest,"Study.ref"])
  
  # Standard deviations, as computed above (actually a delta Ct sd) 
  sx <- filter(qPCR_time_course.summary,BCS==x1 & ExposureTime==timei & Study==study.x)$deltaCt.RNU6.sd
  sy<-filter(qPCR_time_course.summary,BCS==y1 & ExposureTime==timei  & Study==study.y)$deltaCt.RNU6.sd
  
  # deltaCt mean
  mx<-filter(qPCR_time_course.summary,BCS==x1 & ExposureTime==timei & Study==study.x)$deltaCt.RNU6
  my<-filter(qPCR_time_course.summary,BCS==y1 & ExposureTime==timei & Study==study.y)$deltaCt.RNU6
  
  ## Compute t statistic/p value using mean, s.d. and sample size for two groups
  tes<-t.test2(m1 = mx,
               m2= my,
               s1=sx,
               s2=sy,
               n1=3,
               n2=3) 
  
  qpcr_sigtest[sigtest,"pval"]<-tes["p-value"]
  qpcr_sigtest[sigtest,"mean_diff"]<-tes["Difference of means"]
  
  qpcr_sigtest[sigtest,"se"]<-sqrt(sy^2+sx^2)
  
  ## ddCt >0 implies x transcript is less abundant than y and ddCt<0 implies x is more abudnact than y. The relative magnatitude on a plot of 2^(-x) != 2(x) although the interpretability is the same (transcript is either 2^-x less abudnant than y or 2^x more abudnant than y). Therefore, take absolute ddCt before raising it to 2.   
  ##Note ddct is actually 2^-ddct
  
  if(-qpcr_sigtest[sigtest,"mean_diff"]<0){
    
    qpcr_sigtest[sigtest,"ddct"]<-(-2^(tes["Difference of means"]))
    qpcr_sigtest[sigtest,"ddct_se"]<-(-2^(tes["Difference of means"]+
                                            qpcr_sigtest[sigtest,"se"]))
    qpcr_sigtest[sigtest,"ddct_se_m"]<-(-2^(tes["Difference of means"]-
                                              qpcr_sigtest[sigtest,"se"]))
  }else{
    
    qpcr_sigtest[sigtest,"ddct"]<-2^(-tes["Difference of means"]) 
    
    qpcr_sigtest[sigtest,"ddct_se"]<-2^((-tes["Difference of means"])+
                                          qpcr_sigtest[sigtest,"se"])
    qpcr_sigtest[sigtest,"ddct_se_m"]<-2^((-tes["Difference of means"])-
                                            qpcr_sigtest[sigtest,"se"])
    
  }
}

## Annotate sig tests
qpcr_sigtest[qpcr_sigtest$pval<=pval_threshold & !is.na(qpcr_sigtest$pval),"sig"]<-"*"

##convert exposure time to numeric for plot
qpcr_sigtest$ExposureTime.n<-as.numeric(gsub(qpcr_sigtest$ExposureTime,pattern = "hr",replacement = ""))

qpcr_sigtest$ref<-as.character(qpcr_sigtest$reference)
qpcr_sigtest$comp<-as.character(qpcr_sigtest$comparison)

## Assign line types
qpcr_sigtest[qpcr_sigtest$ref=="G. vaginalis","lt"]<-1#"dashed"
qpcr_sigtest[qpcr_sigtest$ref=="Cell Culture Medium","lt"]<-3#"solid"
#cairo_ps(paste0(R_script_output_directory,"miR_long_plot_presentation.eps"),width = 11.5,height = 6)

qpcr_sigtest$comparison<-factor(qpcr_sigtest$comparison,levels = c("L. crispatus", "L. jensenii","L. iners","G. vaginalis", "PH_766","D_10","L_10","DL_10","0_06_D","0_06_L"),ordered = T,labels = c("L. crispatus", "L. jensenii","L. iners","G. vaginalis", "1% lactic acid, pH 7.66","0.1% D-lactic acid","0.1% L-lactic acid","0.1% DL-lactic acid","0.06% D-lactic acid","0.06% L-lactic acid"))

qpcr_sigtest$reference<-factor(qpcr_sigtest$reference,levels = c("L. jensenii","L. iners","G. vaginalis", "Cell Culture Medium", "L_10","DL_10","0_06_D","0_06_L"),ordered = T,labels = c( "L. jensenii","L. iners","G. vaginalis", "Cell Culture Medium","0.1% L-lactic acid","0.1% DL-lactic acid","0.06% D-lactic acid","0.06% L-lactic acid"))


## Plot qPCR Timecourse (put lactic acid comparison in sep. figure)

PLOT_QPCR_TIMECOURSE<-
   ggplot(filter(qpcr_sigtest,reference=="Cell Culture Medium" &
                   comp %in% c("L. crispatus","L. jensenii","L. iners","G. vaginalis","DL_10") ## FIXME sloppy
                 ), 
         aes(x=ExposureTime.n,y=-mean_diff,col=comp))+
  theme_bw()+
    journalFormat+
  geom_hline(yintercept = 0,col="#9e9ac8",size=2.5)+
  theme(plot.margin=unit(c(0,0,0,0),units = "pt"), text=element_text(size=16))+
  geom_line(size=1.5,show.legend = T)+
  geom_errorbar(aes(ymax=-mean_diff+se,ymin=-mean_diff-se),size=1,show.legend = F)+
  geom_point(size=6,show.legend = F)+ 
  ylab(expression(paste(-Delta,Delta,"Ct")))+
  xlab("Exposure Time (hours)")+
  scale_color_manual(name=expression(paste("Lactobacillus ",Delta,"Ct")),values = color_scheme_BCS)+
  scale_x_continuous(breaks=c(0.5,1,4,8,11,13,16,19,22),minor_breaks = NULL)+
  scale_linetype_manual(name=expression(paste("Reference ",Delta,"Ct")),values=as.numeric(qpcr_sigtest$lt))+
  geom_text(aes(label=sig),col='#252525',size=8,nudge_y = -.05)


cairo_ps(paste0(thesis_figures_directory,FIGURE_QPCR_TIMECOURSE),width = 9,height = 6)
PLOT_QPCR_TIMECOURSE
dev.off()

## Show D and L lactic acid qPCR results compared to other references
ddct_LacticAcid_plot<-
  ggplot(filter(qpcr_sigtest,ExposureTime=="4hr" & !qpcr_sigtest$reference=="0.06% D-lactic acid"),aes(x=reference,y=-mean_diff,col=comparison))+
  theme_bw()+
  geom_hline(yintercept = 0,col="#9e9ac8",size=2.5)+
  theme(plot.margin=unit(c(0,0,0,0),units = "pt"), text=element_text(size=14))+
  geom_errorbar(aes(ymax=-mean_diff+se,ymin=-mean_diff-se),width=.2,show.legend = F,position = position_dodge(width = .2))+
  geom_point(size=6,show.legend = T,position = position_dodge(width = .2))+ 
  ylab(expression(paste(-Delta,Delta,"Ct")))+
  xlab(expression(paste("Reference ",Delta,"Ct")))+
  scale_color_manual(name=expression(paste("Exposure ",Delta,"Ct")),values = color_scheme_BCS)+
  geom_text(aes(label=sig),col='#252525',size=8,nudge_y = -.05)

cairo_ps(paste0(thesis_figures_directory,FIGURE_DL_LACTICACID_QPCR),width = 9,height = 6)
ddct_LacticAcid_plot
dev.off()

paste0("Additionally, the Ct of miR-193b expression after 4 hours of exposure to 0.1% D-lactic acid relative to 0.1% L-lactic acid was found to be non-significant ( p=",qpcr_sigtest[qpcr_sigtest$comparison=="0.1% D-lactic acid" & qpcr_sigtest$reference=="0.1% L-lactic acid","pval"],").") 

## Write plot as a table
qpcr_sigtests<-dplyr::select(qpcr_sigtest[with(qpcr_sigtest,order(ExposureTime.n,comparison,reference)),],c(comparison,reference,ExposureTime,mean_diff,pval,sig))
qpcr_sigtests$Figure<-gsub(FIGURE_QPCR_TIMECOURSE,pattern=".eps",replacement = "")

write.csv(qpcr_sigtests,file=paste0(thesis_tables_directory,TABLE_QPCR_TIMECOURSE),row.names = F)

```

## Quantify VK2 proliferation (function of miR-193b) exposed to BCS   

*  Expose VK2 cells to BCS   
*  Measure EdU detection and filled scratch area 

```{r Scratch_EDU,eval=T}


###FIXME this should now be set up closert to qPCR given it is a time course

#write.csv(in_vitro_experiments,"~/Desktop/inv.csv")
## Breadcrumb: 
# In_Vitro_Experiments.Rdata in PreparePipeline/Prepare_EdU_Scratch.Rmd
load(file=paste0(R_script_input_directory,"In_Vitro_Experiments.Rdata")) 
#in_vitro_experiments<-read.csv(paste0("~/smith_thesis_2017/PreparePipeline/","In_vitro_Experiments.csv")) ## pre-compiled csv datatable. 
## Melt scratch assay data for easier handling
scratch.m<-filter(in_vitro_experiments,Experiment=="Scratch" & Observation == "Proliferation" & !grepl(x = as.character(in_vitro_experiments$Treatment),pattern = "CONTROL_")) %>% dplyr::select(-c(Experiment,Coverslip)) 

## Re-name the melted data
names(scratch.m)<-c("Observation","BCS","ExposureTime","percent_cells","Field")


## //////////////
### Scratch Assay
## //////////////

#Factor BCS so that order is enforced
##FIXME some of this can be moved into the prepare script!
scratch.m$BCS<-factor(scratch.m$BCS,ordered = T,levels = c("L. crispatus","L. jensenii","L. iners","G. vaginalis","Cell Culture Medium","0.1% DL lactic acid"))
scratch_sigtest.all<-data.frame(temp="NA")
scratch.summary.all<-data.frame(temp="NA")

for(et in c("1hr","4hr","8hr","13hr","16hr","19hr","22hr")){
##FIXME this is sloppy
  ##Ignore sig tests for now since the plots are clearly sig... too many comparisons
  ## Create a significance testing data frame to hold results

setup_sigtest.data<-setup_sigtest(pval_threshold = pval_threshold,raw_data = dplyr::filter(scratch.m,ExposureTime==et),test_function = "t.test",Experiment = "Scratch")

scratch_sigtest<-setup_sigtest.data$sigtest
scratch_sigtest$ExposureTime<-et
scratch.summary<-setup_sigtest.data$summary_stats
scratch.summary$ExposureTime.n<-as.numeric(gsub(et,pattern = "hr",replacement = ""))

scratch_sigtest.all<-rbind.fill(scratch_sigtest,scratch_sigtest.all)
scratch.summary.all<-rbind.fill(scratch.summary,scratch.summary.all)
}


PLOT_SCRATCH<-
   ggplot(scratch.summary.all,
         aes(x=ExposureTime.n,y=mean,col=BCS))+
  theme_bw()+
  theme(plot.margin=unit(c(0,0,0,0),units = "pt"), text=element_text(size=16))+
  geom_line(size=1.5,show.legend = T)+ 
  geom_errorbar(aes(ymax=mean+sd,ymin=mean-sd),size=1,width=.5,show.legend = F)+
  geom_point(size=6,show.legend = F)+ 
  ylab("Filled Scratch Area (%)")+
    journalFormat+
  xlab("Exposure Time (hours)")+
  scale_color_manual(values = color_scheme_BCS)+
  scale_x_continuous(breaks=c(1,4,8,13,16,19,22),minor_breaks = NULL,limits=c(0,22.5))+
  scale_y_continuous(limits=c(1.05*min(scratch.summary.all$mean-scratch.summary.all$sd,na.rm = T),102))

##Write plot

postscript(paste0(thesis_figures_directory,FIGURE_SCRATCH_QUANT),width = 8,height = 5.5)
plot(PLOT_SCRATCH) 
dev.off()

## //////////////
### EdU Assay
## //////////////

##FIXME - standardize with scratch since they are basically the same plot template

## Melt EdU assay data for easier handling, clean up input
EdU.m<-filter(in_vitro_experiments,Experiment=="Scratch" & Observation == "EdU") %>% dplyr::select(-c(Experiment)) #melt(scratch)
EdU.m<-rename(EdU.m,BCS=Treatment)

#Factor BCS so that order is enforced
EdU.m$BCS<-factor(EdU.m$BCS,ordered = T,levels = c("L. crispatus","L. jensenii","L. iners","G. vaginalis","Cell Culture Medium","0.1% DL lactic acid"))
EdU.m$percent_cells<-EdU.m$percent_cells+abs(rnorm(nrow(EdU.m),1E-3,0.01))
statbars<-EdU_sigtest<-statbars2<-EdU.summary<-data.frame(temp="NA")
for(et in c("1hr","4hr","8hr","13hr","16hr","19hr","22hr")){
## Create a significance testing data frame to hold results
setup_sigtest.data<-setup_sigtest(pval_threshold = pval_threshold,raw_data = dplyr::filter(EdU.m,ExposureTime==et),test_function = "t.test",Experiment = "Scratch")

statbars<-rbind.fill(data.frame(setup_sigtest.data$statbars,ExposureTime=as.numeric(gsub(et,pattern = "hr",replacement = ""))),statbars)
statbars2<-rbind.fill(data.frame(setup_sigtest.data$statbars2,ExposureTime=as.numeric(gsub(et,pattern = "hr",replacement = ""))),statbars2)
EdU_sigtest<-rbind.fill(data.frame(setup_sigtest.data$sigtest,ExposureTime=as.numeric(gsub(et,pattern = "hr",replacement = ""))),EdU_sigtest)
EdU.summary<-rbind.fill(data.frame(setup_sigtest.data$summary_stats,ExposureTime=as.numeric(gsub(et,pattern = "hr",replacement = ""))),EdU.summary)
}
EdU.summary$BCS<-factor(EdU.summary$BCS,rev(unique(EdU.summary$BCS)),ordered = T)
PLOT_EDU<-
   ggplot(EdU.summary,
         aes(x=ExposureTime,y=mean,col=BCS))+
  theme_bw()+
  theme(plot.margin=unit(c(0,0,0,0),units = "pt"), text=element_text(size=16))+
  geom_line(size=1.5,show.legend = T)+ 
  geom_errorbar(aes(ymax=mean+sd,ymin=mean-sd),size=1,width=.5,show.legend = F)+
  geom_point(size=6,show.legend = F)+ 
       journalFormat+
  ylab("Epithelial Cells Positive for EdU (%)")+
  xlab("Exposure Time (hours)")+
  scale_color_manual(values = color_scheme_BCS)+
  scale_x_continuous(breaks=c(1,4,8,13,16,19,22),minor_breaks = NULL,limits=c(0,22.5))+
  scale_y_continuous(limits=c(1.1*min(EdU.summary$mean-EdU.summary$sd,na.rm = T),102))

##Write plot

postscript(paste0(thesis_figures_directory,FIGURE_EDU_QUANT),width = 8,height = 5.5)
plot(PLOT_EDU)
dev.off()

##FIXME

bcs_media_sigtest.all<-data.frame(temp="NA")
bcs_media_summary.all<-data.frame(temp="NA")
bcs_media<-dplyr::filter(in_vitro_experiments,Treatment %in% c("CONTROL_NYCmed_20pct","CONTROL_TSBmed_20pct"))# %>% dplyr::select(-c(Experiment)) 
bcs_media<-rename(bcs_media,BCS=Treatment)

control_meds.sig.migration_p.value<-t.test(dplyr::filter(bcs_media,Observation=="Proliferation" & BCS=="CONTROL_NYCmed_20pct")$percent_cells,
       dplyr::filter(bcs_media,Observation=="Proliferation" & BCS=="CONTROL_TSBmed_20pct")$percent_cells,alternative = "two.sided",paired = F)

control_meds.sig.edu_p.value<-t.test(dplyr::filter(bcs_media,Observation=="EdU" & BCS=="CONTROL_NYCmed_20pct")$percent_cells,
       dplyr::filter(bcs_media,Observation=="EdU" & BCS=="CONTROL_TSBmed_20pct")$percent_cells,alternative = "two.sided",paired = F)

control_meds.summary<-ddply(bcs_media,c("Observation","BCS"),summarise,mean=mean(percent_cells),sd=sd(percent_cells))

NYC.Migration<-dplyr::filter(control_meds.summary,Observation=="Proliferation" & BCS=="CONTROL_NYCmed_20pct")
NYC.EdU<-dplyr::filter(control_meds.summary,Observation=="EdU" & BCS=="CONTROL_NYCmed_20pct")
TSB.Migration<-dplyr::filter(control_meds.summary,Observation=="Proliferation" & BCS=="CONTROL_TSBmed_20pct")
TSB.EdU<-dplyr::filter(control_meds.summary,Observation=="EdU" & BCS=="CONTROL_TSBmed_20pct")

paste0("Percent cell proliferation between 20% NYC-III and TSB bacterial culture media as evaluated by scratch assay and EdU was not significant (scratch area filled p=",
       #(mean +/- standard deviation NYC-III scratch area filled",
       #NYC.Migration$mean,
       #"% +/- ",
       #NYC.Migration$sd,
       #"%, mean +/- standard deviation TSB  cratch area filled=,",
       #TSB.Migration$mean,
       #"% +/- ",
      #  TSB.Migration$sd,
       #"%,
       control_meds.sig.migration_p.value$p.value,
       #"(mean +/- standard deviation NYC-III scratch area filled=",
       #NYC.EdU$mean,
       #"% +/- ",
       #NYC.EdU$sd,
       #"%, mean +/- standard deviation TSB  cratch area filled=,",
       #TSB.EdU$mean,
       #"% +/- ",
       #TSB.EdU$sd,
       #"%, 
      ", percent EdU p=",
       control_meds.sig.edu_p.value$p.value,")")


##FIXME no longer applicable with time course ddata
## Combine proliferation sig tests and write to file
proliferation_sigtests<-rbind(data.frame(Figure=gsub(FIGURE_SCRATCH_QUANT,replacement = "",pattern = ".eps"),scratch_sigtest.all),data.frame(Figure=gsub(FIGURE_EDU_QUANT,replacement = "",pattern = ".eps"),EdU_sigtest))
proliferation_sigtests<-proliferation_sigtests[with(proliferation_sigtests, order(Figure, x,xend)),]
proliferation_sigtests<-dplyr::select(proliferation_sigtests,Figure,xref,reference,ExposureTime,pval,mean_diff,sig)


sig.migration<-t.test(dplyr::filter(control_meds,Observation=="Proliferation" & Treatment=="CONTROL_TSBmed_20pct")$percent_cells,
dplyr::filter(control_meds,Observation=="Proliferation" & Treatment=="CONTROL_NYCmed_20pct")$percent_cells)


proliferation_sigtests<-rbind(proliferation_sigtests,
                              data.frame(Figure="FIGURE_S5_B",
                                         xref="TSB",
                                         reference="NYC",
                                         ExposureTime="13hr",
                              pval=sig.migration$p.value,
                              mean_diff=sig.migration$estimate[2] - sig.migration$estimate[1],
                              sig=ifelse(sig.migration$p.value<pval_threshold,"*","N.S.")))
 

sig.edu<-t.test(dplyr::filter(control_meds,Observation=="EdU" & Treatment=="CONTROL_TSBmed_20pct")$percent_cells,
dplyr::filter(control_meds,Observation=="EdU" & Treatment=="CONTROL_NYCmed_20pct")$percent_cells)


proliferation_sigtests<-rbind(proliferation_sigtests,
                              data.frame(Figure="FIGURE_S5_C",
                                         xref="TSB",
                                         reference="NYC",
                                         ExposureTime="13hr",
                              pval=sig.edu$p.value,
                              mean_diff=sig.edu$estimate[2] - sig.edu$estimate[1],
                              sig=ifelse(sig.edu$p.value<pval_threshold,"*","N.S.")))
 
##FIXME- rename proliferation to migration and EdU to proliferation
control_meds.summary$Observation<-factor(control_meds.summary$Observation,levels = c("Proliferation","EdU"),labels=c("Migration (scratch)","Proliferation (EdU)"),ordered = T)

PLOT_nyc_v_tsb<-ggplot(control_meds.summary)+
  geom_bar(aes(x=BCS,y=mean,fill=BCS),stat="identity",show.legend =F)+
  geom_errorbar(aes(x=BCS,ymax=mean+sd,ymin=mean-sd),width=.2)+
  journalFormat+ylab("Percent scratch area filled | Percent cells EdU")+
  xlab("VK2 Cell Culture Medium + 20% Bacterial Culture Medium")+
  scale_y_continuous(limits=c(0,100))+
  annotate("text",x=1.5,y=90,label="N.S.",size=4)+
  geom_segment(data=data.frame(x=1,xend=2,y=87,yend=87),aes(x=x,xend=xend,y=y,yend=yend),size=1)+
  geom_segment(data=data.frame(x=c(1,2),xend=c(1,2),y=c(85,85),yend=c(87,87)),aes(x=x,xend=xend,y=y,yend=yend),size=1)+
  facet_wrap(~Observation,nrow=1)

postscript(file=paste0(thesis_figures_directory,FIGURE_nyc_v_tsb),width = 8,height = 5.5)
PLOT_nyc_v_tsb
dev.off()

write.csv(proliferation_sigtests,file=paste0(thesis_tables_directory,TABLE_EDU_SCRATCH_QUANT),row.names = F)

## PErcent proliferation in 1% DL lactic acid, ph buffered 7.66 expoed cells. Easier to hardcode than read in table
DL_PH766_percent_proliferation<-data.frame(DL_1pct_766=c(
7.706708323,
4.631917974,
8.766697901,
7.771365876)
)
mean(DL_PH766_percent_proliferation$DL_1pct_766)
sd(DL_PH766_percent_proliferation$DL_1pct_766)
t.test(x = scratch.m[scratch.m$BCS=="Cell Culture Medium","percent_cells"],y = DL_PH766_percent_proliferation$DL_1pct_766)
t.test(x = scratch.m[scratch.m$BCS=="G. vaginalis","percent_cells"],y = DL_PH766_percent_proliferation$DL_1pct_766)


```

## Create bar plot for CCND1 protein expression quant (Western blot)
```{r Western,eval=T}

load(paste0(R_script_input_directory,"Western.Rmd"))

western.plot<-ggplot(dplyr::filter(Western,!ExposureTime==24),aes(x=as.factor(ExposureTime),y=BCS_MEDIA,fill=BCS))+geom_bar(stat = "identity",width=.75, position = "dodge")+
  scale_fill_manual(values=color_scheme_BCS)+
  journalFormat+
  ylab("Normalized CCND1 relative to Cell Culture Media")+
  xlab("Exposure time (hours)")


postscript(file=paste0(thesis_figures_directory,FIGURE_WESTERN_QUANT),width = 8,height = 5.5)
plot(western.plot)
dev.off()

```

## Inhbit A2EN cell proliferation & Observe CT Infectivity  

*  Expose A2EN cells to proliferation inhibitors   
*  Infect w/ CT  
*  Measure  EdU detection and infection 

```{r CT_Infection,eval=T}
load(file=paste0(R_script_input_directory,"In_Vitro_Experiments.Rdata"))

## Melt Ct infection assay data for easier handling
pi.m<-filter(in_vitro_experiments,Experiment=="Infection") %>% dplyr::select(-Experiment) 

## Factor Treatment to enforce order and make labels
pi.m$Treatment<-factor(pi.m$Treatment,levels = c("Cdk4 400nM","Fascaplysin 350nM","Cell Culture Medium"),ordered = T,labels = c("CAS 546102-60-7","Fascaplysin","Cell Culture Medium"))

## Create a significance testing data frame to hold results
setup_sigtest.data<-setup_sigtest(pval_threshold = pval_threshold,raw_data = pi.m,test_function = "t.test2",Experiment = "Infection")

pi.summary<-setup_sigtest.data$summary_stats
ct_sigtest<-setup_sigtest.data$sigtest
statbars<-setup_sigtest.data$statbars
statbars2<-setup_sigtest.data$statbars2

## Split data into infection and proliferation (EdU) data
ct_sigtest_inf<-filter(ct_sigtest,Observation=="Infectivity")
ct_sigtest_pro<-filter(ct_sigtest,Observation=="Proliferation")

## Plot Infectivity
Fig3_inf<-ggplot(filter(pi.summary,Observation=="Infectivity"),aes(x=Treatment,y=grand_mean))+geom_bar(aes(fill=Treatment),stat="identity",show.legend = F)+geom_errorbar(aes(ymin=grand_mean-grand_sd,ymax=grand_mean+grand_sd,width=.3))+
  scale_fill_manual(values=c("Cell Culture Medium"='blue',"Fascaplysin"="#fdb863","CAS 546102-60-7"="#b2df8a"))+
  journalFormat+
  theme(plot.margin=unit(c(0,0,0,0),units = "pt"), text=element_text(size=16),axis.text.x=element_text(angle = 45,vjust = 1,hjust = 1,size=20))+scale_y_continuous(limits=c(0,102))+
  annotate("text",x=ct_sigtest_inf$midpoints,y=ct_sigtest_inf$y.sig,label=ct_sigtest_inf$sig,size=2.5)+
  geom_segment(data=statbars,aes(x=x,xend=xend,y=y,yend=yend))+
  geom_segment(data=statbars2,aes(x=x,xend=xend,y=y,yend=yend),col="Black")+ylab("Epithelial Cells with C. trachomatis Inclusion (%)")

## Write plot


postscript(paste0(thesis_figures_directory,FIGURE_CT_INFECT_QUANT),width=6,height = 5.5)#,width = 8,height = 5.5)
plot(Fig3_inf)
dev.off()

##Plot Ct Assay EdU

Fig3_prolif<-ggplot(filter(pi.summary,Observation=="Proliferation"),aes(x=Treatment,y=grand_mean))+geom_bar(stat="identity",show.legend = F,aes(fill=Treatment))+geom_errorbar(aes(ymin=grand_mean-grand_sd,ymax=grand_mean+grand_sd,width=.3))+journalFormat+scale_fill_manual(values=c("Cell Culture Medium"='blue',SIS3="#c2a5cf","Fascaplysin"="#fdb863","CAS 546102-60-7"="#b2df8a"))+ylab("Epithelial Cells Positive for EdU (%)")+
  theme(plot.margin=unit(c(0,0,0,0),units = "pt"), text=element_text(size=20),axis.text.x=element_text(angle = 45,vjust = 1,hjust = 1))+scale_y_continuous(limits=c(0,102))+
  annotate("text",x=ct_sigtest_pro$midpoints,y=ct_sigtest_pro$y.sig,label=ct_sigtest_pro$sig,size=2.5)+
  geom_segment(data=statbars,aes(x=x,xend=xend,y=y,yend=yend))+
  geom_segment(data=statbars2,aes(x=x,xend=xend,y=y,yend=yend),col="Black")

##Write plot
postscript(paste0(thesis_figures_directory,FIGURE_CT_EDU_QUANT),width=6,height = 5.5)#width = 8,height = 5.5)
plot(Fig3_prolif)
dev.off()

## Print text describing results

## Create variables to hold values for printing
ct_sigtest<-unique(ct_sigtest)
pro_C<-ct_sigtest[ct_sigtest$Observation=="Proliferation" & ct_sigtest$reference=="Cell Culture Medium" & ct_sigtest$xref=="CAS 546102-60-7" ,c("mean_diff","pval")]
pro_F<-ct_sigtest[ct_sigtest$Observation=="Proliferation" & ct_sigtest$reference=="Cell Culture Medium" & ct_sigtest$xref=="Fascaplysin" ,c("mean_diff","pval")]

ct_C<-ct_sigtest[ct_sigtest$Observation=="Infectivity" & ct_sigtest$reference=="Cell Culture Medium" & ct_sigtest$xref=="CAS 546102-60-7" ,c("mean_diff","pval")]
ct_F<-ct_sigtest[ct_sigtest$Observation=="Infectivity" & ct_sigtest$reference=="Cell Culture Medium" & ct_sigtest$xref=="Fascaplysin" ,c("mean_diff","pval")]


paste0("Epithelial cell proliferation was decreased by ",-pro_C$mean_diff,"% (p=",pro_C$pval,") in CAS 546102-60-7 and ",-pro_F$mean_diff,"% (p=",pro_F$pval,") in Fascaplysin treated cells relative to Cell Culture Medium treated cells, respectively")

paste0("C. trachomatis infection was decreased by ",-ct_C$mean_diff,"% (p=",ct_C$pval,") in CAS 546102-60-7 and ",-ct_F$mean_diff,"% (p=",ct_F$pval,") in Fascaplysin treated cells relative to Cell Culture Medium treated cells, respectively")

## Calcualte correlation between mean infectivity and mean proliferation
corr_mx<-spread(pi.m,Observation, value=percent_cells)
ggplot(corr_mx,aes(y=Infectivity,x=Proliferation,col=Treatment))+geom_point()+ggtitle("Ct Infectivity vs EdU")+journalFormat+ylab("% Infectivity")+xlab("% EdU")+geom_point(data=spread(dplyr::select(pi.summary,-c(grand_sd,n)),Observation,grand_mean),aes(y=Infectivity,x=Proliferation,col=Treatment),size=10,pch=2)

ct_pro_lm<-lm(formula = Infectivity~Proliferation,data = data.frame(corr_mx[!rowSums(is.na(corr_mx))>0,]))

ct_pro_summary_lm<-lm(Infectivity~Proliferation,data=spread(dplyr::select(pi.summary,-c(grand_sd,n)),Observation,grand_mean))

summary(ct_pro_lm)
summary(ct_pro_summary_lm)



## Combine proliferation sig tests and write to file
ct_sigtests<-data.frame(Figure=paste0("FIGURE_4",ct_sigtest$Observation),ct_sigtest)
ct_sigtests<-dplyr::select(ct_sigtests[with(ct_sigtests,order(Observation,x,xend)),],c(Figure,xref,reference,pval,mean_diff,sig))

write.csv(ct_sigtests,file=paste0(thesis_tables_directory,TABLE_CT_QUANT),row.names=F,quote=F)

```

# Subject Longitudinal Plots (Figure 1)
Create longitudinal plots for subjects used in study

```{r plot_subject_figures,eval=T}
### Load previously prepared 16S metataxonomic data. metatdata
load(file=paste0(R_script_input_directory,"subject_plot_data.Rdata"))
OTU_METADATA<-subject_plot_data$OTU_METADATA ## metadata
rRNA_16S<-subject_plot_data$relativeAbundance ## taxa assignments/relative abundances
sampleInfoColNames<-subject_plot_data$sampleInfoColNames ## holds which column names are associated with metadata

miRNA_extractions<-SRL_meta_table[is.na(SRL_meta_table$QC_removal_stage) & !is.na(SRL_meta_table$CST),] ## Samples associated with miRNA libraries used in final study- i.e., don't include samples removed due to poor QC. 
## Also adding in only samples that have 16S data as well. Will wind up dropping 16 samples, but it is what it is... 

subject_plot_list<- unique(miRNA_extractions$SID) ## Subject IDs to plot

global_species_list<-NULL ## Initialize a container to store all species plotted in fig 1 as a legend

for(s in subject_plot_list){  ## iterate through the subject's
  print(s)
  #s<-"UAB008"
  ## ////////////////////////////////// ##
  ## ///  Subeset Data by Subject   /// ##
  ## ////////////////////////////////// ##
  
  relabundance<-filter(rRNA_16S,SID==s)  ## taxa relative abundance for subject
  otu_count<-OTU_METADATA[OTU_METADATA$SID==s,] ## Metadata for subject
  
  ## Drop 16S samples that have less than the threshold for high confidence taxa assignments
  lowCountThreshold<-1000
  low_count_samples<-otu_count[otu_count$X16S_total_counts<lowCountThreshold,"SERIAL"]
  
  ##Nugent score data for subject
  nugent<-filter(OTU_METADATA,SID==s) %>% dplyr::select(SERIAL,NUGENT_SCORE)
  nugent$color<-"grey"
  ## Colors/handling for missing Nugent scores
  nugent[is.na(nugent$NUGENT_SCORE),"color"]<-"red"
  nugent[!is.na(nugent$NUGENT_SCORE),"color"]<-"black"
  nugent[is.na(nugent$NUGENT_SCORE),"NUGENT_SCORE"]<-(-1)
  
  ## pH data for subject
  ph<-filter(OTU_METADATA,SID==s) %>% dplyr::select(SERIAL,PH)
  ## Colors/handling for missing pH values
  ph$color<-"grey"
  ph[is.na(ph$PH),"color"]<-"red"
  ph[!is.na(ph$PH),"color"]<-"black"
  ph[is.na(ph$PH),"PH"]<-2
  
  ##MEtadata for subject
  dailyDiaryMetadata<-filter(OTU_METADATA,SID==s)
  
  ## ////////////////////////////////// ##
  ##  Determine global plot time scale  ##
  ## ////////////////////////////////// ##
  
  SERIAL_global<-sort(unique(dailyDiaryMetadata$SERIAL,relabundance$SERIAL,miRNA_extractions$SERIAL)) ## All time points
  day<-SERIAL_global%%7 ## Day relative to all SERIALized time points
  day[SERIAL_global%%7==0]<-7 ## the mod calcualtion causes all day 7 to be 0. Repalce w/ 7. 
  week<-((SERIAL_global-day)/7)+1 ## Week back calculated from SERIALized time points. 
  time_points<-data.frame(SERIAL=SERIAL_global, day=day, week=week) ## container
  time_points$plot_label=""
  time_points[time_points$day==7,"plot_label"]<-time_points[time_points$day==7,"week"] ## plot labels every week
  global_min_time<-max(c(min(miRNA_extractions[miRNA_extractions$SID==s,"SERIAL"]-5),min(SERIAL_global,na.rm = T) ))
  
  ## min time point -  the max of the min of either miR extractions or all data
  global_max_time<-min(c(max(miRNA_extractions[miRNA_extractions$SID==s,"SERIAL"]+5),max(SERIAL_global,na.rm = T))) ## max time point across all data
  removed_samples
  ## cleaner way to define as variable for X axis:
  time_breaks<-time_points$SERIAL 
  time_label<-time_points$plot_label
  time_limits<-c(global_min_time-1,global_max_time+1)
  
  ## Determine the min & max of time points
  rect_min<-time_breaks[!time_breaks%in% miRNA_extractions[miRNA_extractions$SID==s,"SERIAL"]]-.5
  rect_max<-time_breaks[!time_breaks%in% miRNA_extractions[miRNA_extractions$SID==s,"SERIAL"]]+.5
  
  ## ////////////////////////////////// ##
  ## / Determine most adbundant species ##
  ## ////////////////////////////////// ##
  
  # ////////////////////////////////////////////////////////////////////////
  # Most abundant species defined per subject & based on cutoff. ##
  # All other taxa binned into "other" #
  # ////////////////////////////////////////////////////////////////////////
  
  ### Grab just relative abundances, no sample info columns:
  relabundance[is.na(relabundance)]<-0
  relabundance<-relabundance[relabundance$SERIAL>=time_limits[1] &  relabundance$SERIAL<=time_limits[2],]
  relabundance_for_max_calc<-relabundance[,!(names(relabundance) %in% sampleInfoColNames)]
  
  ### Calucalte max for each taxa
  max_relabundances<-apply(relabundance_for_max_calc,2,max)
  ### number of species whose rel abundance is above a certain threshold
  numHighAbundSpecies<-sum(max_relabundances>raThreshold,na.rm = T) 
  
  ##### Plot either the top X most abundant species, 
  #####   or the most abundant species above max plottable species, whichever is lesser.
  #####   This helps prevent "taxa overload" on the plot. 
  
  most_abundant_species<-"" ## Will hold names of most abundant species. 
  if(numHighAbundSpecies>nSpecies){ 
    most_abundant_species<-names(sort(max_relabundances,decreasing = T))[1:nSpecies]
  }else{
    most_abundant_species<- names(max_relabundances[!is.na(max_relabundances) & max_relabundances>raThreshold])
  }
  
  ##### Pull the most abundant species (defined above), bin the remainder into "other".
  #####   Also recombine the 'sample info' onto most abundant/other species table. 
  
  most_abundant_relabundance<-relabundance[most_abundant_species]
  other_relabundance<-rowSums(relabundance[!(names(relabundance) %in% most_abundant_species | 
                                               names(relabundance) %in% sampleInfoColNames)],na.rm=T)
  otu_count_relabundance<-cbind(relabundance[names(relabundance) %in% c("SERIAL",sampleInfoColNames)],
                                most_abundant_relabundance,Other=other_relabundance)
  
  ##Update global species list with any new species
  global_species_list<-unique(c(most_abundant_species,global_species_list,"Other"))
  
  
  reshape_names<-c("Pre_QC_ID","SID","UID","SERIAL")
  otu_count_reshape<-melt(data = otu_count_relabundance,id.vars=reshape_names)
  names(otu_count_reshape)<-c(reshape_names,"species","count") 
  
  ### ///////////////////////////////////////////// ###
  ### /////////         OTU Plot          ///////// ###
  ### ///////////////////////////////////////////// ###
  
  otuPlot<- ggplot(otu_count_reshape)+
    geom_area(aes(x=SERIAL,y=count,fill=species),
              stat="identity",show.legend=F,position="fill")+#,width=1)+
    journalFormat+
    theme(axis.ticks = element_blank(),
          axis.title.x=element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = rel(1.5)),
          legend.key.size=unit(8, "points"),
          legend.title = element_blank(),
          legend.text = element_text(size = rel(.5),face="italic"),
          axis.title = element_text(size = rel(sizes)),
          plot.margin=unit(c(2.5,40,2.5,10),units="points"),#margins
          panel.grid.major.y=element_line(colour = "grey73"),
          panel.grid.minor.x  = element_blank())+
    ylab("Phylotype Relative\nAbundance (%)")+
    geom_vline(xintercept = otu_count_reshape$SERIAL,size=rel(.2),col="grey")+
    scale_fill_manual(values=subject_long_taxa_colors)+
    ggtitle(paste0(s))+
    annotate("rect", xmin=rect_min, xmax=rect_max, ymin=0, ymax=1, alpha=alpha_rect, fill=rect_fill)+
    scale_x_continuous(breaks=time_breaks,label=time_label,limits=time_limits)
  
  ## Determine any dropped sample (post QC) time points and place an * above 
  dropped_samples.serial<-otu_count[otu_count$Pre_QC_ID %in% removed_samples$Pre_QC_ID,"SERIAL"]
  if(length(dropped_samples.serial)!=0){otuPlot<-otuPlot+annotate("text", x =dropped_samples.serial , y = 1.01, label = "*",size=8)}
  
  
  ### ///////////////////////////////////////////// ###
  ### /////////      Nugent Plot          ///////// ###
  ### ///////////////////////////////////////////// ###
  nugentPlot<-ggplot(data=nugent)+
    geom_bar(aes(x=as.numeric(SERIAL),y=NUGENT_SCORE,width=.9,fill=color),
             stat="identity",position = position_dodge(width=0.5))+
    journalFormat+
    geom_hline(yintercept = c(3,7),size=rel(1),col="pink")+ ## Defines Nugent score 3 & 7 (BV 
    theme(axis.ticks = element_blank(),
          axis.title.x=element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = rel(1.5*sizes)),
          legend.position="none",
          plot.margin=margins,
          panel.grid.major.y=element_line(colour = "grey73"),
          panel.grid.minor.x  = element_blank())+
    ylab("Nugent\nScore")+
    scale_fill_manual(values=c('black','red'))+
    scale_x_continuous(breaks=time_breaks,label=time_label,limits=time_limits)+
    annotate("rect", xmin=rect_min, xmax=rect_max, ymin=0, ymax=10, alpha=alpha_rect, fill=rect_fill)+
    scale_y_continuous(breaks=c(0,3,7,10),limits=c(-1.25,10.5))
  
  
  ### ///////////////////////////////////////////// ###
  ### /////////     Metadata Plot         ///////// ###
  ### ///////////////////////////////////////////// ###
  metaPlot<-ggplot(dailyDiaryMetadata, aes(x=as.numeric(SERIAL)))+
    geom_point(aes(y=1*(as.numeric(VAG_DIS)==1),size=2),
               pch=16,col='blue',position = position_dodge(width=0.5))+
    geom_point(aes(y=2*(as.numeric(VAG_ODOR)==1),size=2),
               pch=16,col='blue',position = position_dodge(width=0.5))+
    geom_point(aes(y=3*(as.numeric(MENSTRUATION)>0),size=as.numeric(MENSTRUATION)),
               col='red',pch=16, position = position_dodge(width=0.5))+
    scale_size(range = c(2+2,4+2))+ ## for menstru point sizes. Bounded by 3 points on a scale. 
    journalFormat+
    theme(legend.position="none",
          plot.title = element_text(size = rel(sizes)),
          axis.text = element_text(size = rel(2*sizes)),
          axis.title = element_text(size = rel(sizes)),
          axis.title.y=element_text(vjust=.2),
          axis.title.x=element_text(vjust=-.2),
          axis.text.y=element_text(size=rel(0.75)), ## change back to 0.75
          plot.margin=unit(c(-2.5,40,5,5),units="points"),
          panel.grid.major.y=element_line(colour = "grey73"),
          panel.grid.minor.x  = element_blank())+
    xlab("Week")+
    ylab("")+
    annotate("rect", xmin=rect_min, xmax=rect_max, ymin=1, ymax=3, alpha=alpha_rect, fill=rect_fill)+
    scale_x_continuous(breaks=time_breaks,label=time_label,limits=time_limits)+
    scale_y_continuous(breaks=1:3,labels = c("Discharge",
                                             "Odor",
                                             "Menstruation"),limits=c(0.5,3.5))
  
  ###### Determine if subject has low qualiyt daily dairy flag and annotate plot:
  
  if(!sum(is.na(dailyDiaryMetadata$Diary_QUALITY_FLAG))){
    min<-min(timeTable$SERIAL,na.rm = T)
    max<-max(timeTable$SERIAL,na.rm = T)
    middle.x<-(max-min)/2
    middle.y<<-(16-1)/2
    metaPlot<-metaPlot + annotate("text", x = middle.x, 
                                  y = middle.y, label = "?",
                                  size=rel(40),
                                  col="grey")
  }
  
  ### ///////////////////////////////////////////// ###
  ### /////////          pH Plot          ///////// ###
  ### ///////////////////////////////////////////// ###
  ph_normalization_factor<-3.5
  ## Notice the pH value scale is "normalized" by subtracting "ph_normalization_factor" from the actual pH value, then re-labeling the y axis. This is very dangerous, but ggplot will not permit bar plots that start from > 0 . 
  ph[ph$PH==2,"PH"]<-ph_normalization_factor-0.5
  
  phPlot<-ggplot(ph,aes(x=as.numeric(SERIAL),
                        y=as.numeric(PH)-ph_normalization_factor,width=.9,fill=color))+ 
    geom_bar(stat="identity",position = position_dodge(width=0.5))+
    scale_fill_manual(values=c('black','red'))+
    journalFormat+
    geom_hline(yintercept = c(4.5-ph_normalization_factor),
               size=rel(1),col="pink")+ ## Vaginal pH>4.5 one criteria for BV. 
    theme(axis.ticks = element_blank(),
          axis.title.x=element_blank(),
          axis.text.x = element_blank(),
          axis.text = element_text(size = rel(1.5*sizes)),
          legend.position="none",
          plot.margin=margins,
          panel.grid.major.y=element_line(colour = "grey73"),
          panel.grid.minor.x  = element_blank())+
    ylab("pH")+
    annotate("rect", xmin=rect_min, xmax=rect_max, ymin=min(0,min(ph$PH,na.rm = T)-ph_normalization_factor-0.25), ymax= max(ph$PH,na.rm = T)+0.25-ph_normalization_factor, alpha=alpha_rect, fill=rect_fill)+
    scale_x_continuous(breaks=time_breaks,label=time_label,limits=time_limits)+
    scale_y_continuous(breaks=c(4,4.5,5,5.5)-ph_normalization_factor, limits=c(-0.75,5.8+0.25-ph_normalization_factor),labels=c("4","4.5","5","5.5"))
  
  
  ### ///////////////////////////////////////////// ###
  ### /////////    Tie Plots Together     ///////// ###
  ### ///////////////////////////////////////////// ###
  
  ## ////////////////////////// ##
  ## //Define plots as Grobs/// ##
  ## ////////////////////////// ##
  
  grob.otuPlot <- ggplotGrob(otuPlot)
  grob.nugentPlot <- ggplotGrob(nugentPlot)
  grob.phPlot<-ggplotGrob(phPlot)
  grob.metaPlot <- ggplotGrob(metaPlot)
  
  ## ////////////////////////// ##
  ## ///   Find max width   /// ##
  ## ////////////////////////// ##
  maxWidth = grid::unit.pmax(grob.otuPlot$widths[1:6],
                             grob.nugentPlot$widths[1:5],
                             grob.phPlot$widths[1:5],
                             grob.metaPlot$widths[1:5])
  
  ## ////////////////////////// ##
  ## /Redefine common max width ##
  ## ////////////////////////// ##
  grob.nugentPlot$widths[1:6] <- as.list(maxWidth)
  grob.otuPlot$widths[1:6] <- as.list(maxWidth)
  grob.metaPlot$widths[1:6] <- as.list(maxWidth)
  grob.phPlot$widths[1:6]<-as.list(maxWidth)
  
  ### ///////////////////////////////////////////// ###
  ### /////////      Write/Draw Plot     ///////// ###
  ### ///////////////////////////////////////////// ###
  
  cairo_ps(paste0(thesis_figures_directory,FIGURE_SUBJECT_PLOTS,s,".eps"),width = 11,height = 8.5)
  
  grid.arrange(grob.otuPlot,
               grob.nugentPlot,
               grob.phPlot,
               grob.metaPlot, 
               ncol=1,nrow=4,
               heights=c(2.5,1,1,1)) 
  dev.off()
}

## //////
## The following plots the figure legend containing colors for all taxa plotted in for loos

dummy_globalSpeciesList<-data.frame(SERIAL=1,species=global_species_list,count=1)
dummy_globalSpeciesList_plot<-ggplot(dummy_globalSpeciesList)+
  geom_area(aes(x=SERIAL,y=count,fill=species,order=order(as.numeric(as.factor(dummy_globalSpeciesList$species)),decreasing = F)),
            stat="identity",show.legend=T,position="fill")+#,width=1)+
  journalFormat+
  theme(axis.ticks = element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_blank(),
        legend.key.size=unit(8, "points"),
        legend.title = element_blank(),
        legend.text = element_text(size = rel(.5),face="italic"),
        axis.title = element_text(size = rel(sizes)),
        plot.margin=unit(c(2.5,40,2.5,10),units="points"),#margins
        panel.grid.major.y=element_line(colour = "grey73"))+
  scale_fill_manual(values=subject_long_taxa_colors)+
  ggtitle("Global Species List Color Codes")

##Write legend to file
cairo_ps(paste0(thesis_figures_directory,FIGURE_SUBJECT_PLOTS,"GlobalSpeciesList.eps"),width = 11,height = 8.5)
dummy_globalSpeciesList_plot
dev.off()

```

# Ribo-reduced RNA-seq Analysis

```{r RNA_quality_quant}
## RIN Quality Distribition 
TRL_RNA_Sample_QuantQual<-read.csv(paste0(R_script_input_directory,"TRL_RNA_Sample_Quality.csv"),header = F)
#postscript(paste0(thesis_figures_directory,FIGURE_TRL_RIN_HIST))
p<-ggplot(TRL_RNA_Sample_QuantQual,aes(x=V1))+geom_histogram()+xlab("RINe")+ylab("Number of Samples")+geom_vline(xintercept = median(TRL_RNA_Sample_QuantQual$V1),col='red')+journalFormat
plot(p)
dev.off()
plot(p)
summary(TRL_RNA_Sample_QuantQual$V1)
summary(10*TRL_RNA_Sample_QuantQual[TRL_RNA_Sample_QuantQual$V1!=3.1,"V2"])
```

## Read counts data in from server   


```{r connect_unix,eval=FALSE}
### ---------------------------------
## Set up unix connection using sshfs
### ---------------------------------
Sys.which('bash')
Sys.which('sh')
#echo hello world
# 
system("sshfs stsmith@medusa.igs.umaryland.edu:/local/projects-t2/HRBV/ ~/IGS/sshfs_medusa/", intern = FALSE,
       ignore.stdout = FALSE, ignore.stderr = FALSE,
       wait = TRUE, input = NULL, show.output.on.console = TRUE,
       minimized = FALSE, invisible = F)

manifest<-read.table("~/IGS/sshfs_medusa/Config/W100083533.manifest",sep = "\t",header = T)
TRL_counts_table<-data.frame(Feature="None")

for(i in 1:length(manifest$Pre_QC_ID)){
  #i<-3
  sample_name<-as.character(manifest[i,"Pre_QC_ID"])
  sample_path<-paste0("~/IGS/sshfs_medusa/TRL/alignment/TRL_",sample_name,"_aln_hg19/TRL_",sample_name,".hg_counts")
  new_table<-read.table(sample_path,header=T,sep="\t")
  names(new_table)<-c("Feature",sample_name)
  #sub_feature_set<-!new_table$Feature %in% c(filtered_out_features,"mirna_info")
  ## Don't forget about ambigous/none features!
  #new_table<-new_table[sub_feature_set,]
  TRL_counts_table<-merge(TRL_counts_table,new_table,by.x = "Feature",by.y="Feature",all=T)
}

row.names(TRL_counts_table)<-TRL_counts_table$Feature
TRL_counts_table<-TRL_counts_table[!TRL_counts_table$Feature=="None",]
TRL_counts_table<-dplyr::select(TRL_counts_table,-c(Feature,POSITIVE_CONTROL))
head(TRL_counts_table)
write.csv(TRL_counts_table,file=paste0(R_script_output_directory,TABLE_TRL_COUNTS_RAW),row.names = T,quote=F)

```


## Create and Prepare Tables   


```{r create_tables}
TRL_counts_table<-read.csv(file=paste0(R_script_output_directory,TABLE_TRL_COUNTS_RAW),row.names = 1)
write.csv(TRL_counts_table,file=paste0(thesis_tables_directory,TABLE_TRL_COUNTS_RAW),row.names = T,quote=F)

## Filter out ambiguous, no feature reads. Make note of how many there are
TRL_counts_table[row.names(TRL_counts_table) %in% c("alignment_not_unique","no_feature","ambiguous"),]

## Percentage of ambiguous/no feature/non unique reads & alignment stats
(ambig_nofeat<-colSums(TRL_counts_table[row.names(TRL_counts_table) %in% c("alignment_not_unique","no_feature","ambiguous"),]))
(aligned<-colSums(TRL_counts_table[!row.names(TRL_counts_table) %in% c("alignment_not_unique","no_feature","ambiguous"),]))
alignment_stats<-data.frame(aligne=aligned,non_aligned=ambig_nofeat,aligned.percent=100*aligned/colSums(TRL_counts_table))
write.csv(alignment_stats,file = paste0(thesis_tables_directory,TABLE_TRL_ALIGNSTATS))

## Look at proportion of Ambigous/no feature/non-unique
hist(alignment_stats$aligned.percent,main="Percentage of Aligned reads",xlab="Percentage of total trimmed reads",ylab="Proportion")

ambig_nofeat_readpercent.high<-alignment_stats[(100-alignment_stats$aligned.percent)>80,]
paste0("The following samples have ambigous reads >80%: ", str_c(row.names(ambig_nofeat_readpercent.high),collapse = ", "))

#Filter out ambigous reads from counts
TRL_counts_table<-TRL_counts_table[!row.names(TRL_counts_table) %in% c("alignment_not_unique","no_feature","ambiguous"),]
TRL_counts_table<-TRL_counts_table[,order(names(TRL_counts_table))]
hist(log(as.matrix(TRL_counts_table)),main="log10 read counts, all samples",xlab="Log10 read count")

##Create Design Table
TRL_design<-data.frame(Sample=names(TRL_counts_table))
row.names(TRL_design)<-TRL_design$Sample
TRL_design<-separate(TRL_design,Sample,sep = "_",into = c("CellLine","BCS","DROP","ExposureTime","rep")) %>% dplyr::select(-c(CellLine,DROP))

## Create replicate #, lactobacillus indicator. Then combine BCS and exposure time for 'group1' for eventually creating constrasts
TRL_design$rep<-as.numeric(gsub(TRL_design$rep,pattern = "rep",replacement = ""))
TRL_design$Lactobacillus<-grepl(TRL_design$BCS,pattern = "^L")
TRL_design$group1 <- factor(paste(TRL_design$BCS,TRL_design$ExposureTime,sep="."))
TRL_design$BCS<-factor(TRL_design$BCS,levels = c("LCRISPATUS","LJENSENII","LINERS","GVAGINALIS","MEDIA"),ordered = T)
TRL_design$ExposureTime<-factor(TRL_design$ExposureTime,levels = c("4HR","13HR","22HR"),ordered=T)
sample_order<-with(TRL_design,order(ExposureTime,BCS,rep))

##Make an expression set object for coupled handling of coutns and design matrix
TRL_design<-TRL_design[sample_order,]
TRL_counts_table<-TRL_counts_table[,sample_order]
TRL_counts_meta<-ExpressionSet(assayData = as.matrix(TRL_counts_table),phenoData  = AnnotatedDataFrame(TRL_design))

```

## Plot Replicates
```{r plot_reps, eval=F}

## THis may take a while given there are ~20,000 points to plot for each comparison. Consdier trimming the lower expresed reads by using rmlow = log(10,10) for example
postscript(paste0(R_script_output_directory,"TRL_RNASeq_Replicate_plots.eps"),width = 10,height = 8)
hist(colSums(exprs(TRL_counts_meta)))

plot_replicates(TRL_counts_meta,BCS.selection = c("LCRISPATUS"),ExposureTime.selection = c("4HR","13HR","22HR"),logt=T)
plot_replicates(TRL_counts_meta,BCS.selection = c("LJENSENII"),ExposureTime.selection = c("4HR","13HR","22HR"),logt=T)
plot_replicates(TRL_counts_meta,BCS.selection = c("LINERS"),ExposureTime.selection = c("4HR","13HR","22HR"),logt=T)
plot_replicates(TRL_counts_meta,BCS.selection = c("GVAGINALIS"),ExposureTime.selection = c("4HR","13HR","22HR"),logt=T)
plot_replicates(TRL_counts_meta,BCS.selection = c("MEDIA"),ExposureTime.selection = c("4HR","13HR","22HR"),logt=T)

dev.off()
```
## Drop samples
```{r drop_samples}
dropped_TRL_samples<-c("VK2_MEDIA_BCS_4HR_rep3","VK2_LCRISPATUS_BCS_13HR_rep3") ## determined from replciate plots and # of ambigous (rRNA) reads.

##Drop poor QC samples
TRL_counts_meta.qc<-subset_ExpressionSet(TRL_counts_meta,filterOut =dropped_TRL_samples )

## Proportion of genes with at least 1 read across all samples
sum(rowSums(exprs(TRL_counts_meta.qc)>0)>0)/nrow(exprs(TRL_counts_meta.qc))

#exprs(TRL_counts_meta.qc)[rowSums(exprs(TRL_counts_meta.qc)>0)>0,]

##Summary of # of samples with at least one read across all samples
summary(rowSums(exprs(TRL_counts_meta.qc)[rowSums(exprs(TRL_counts_meta.qc)>0)>0,]>0))

##Total post QC samples:
ncol(exprs(TRL_counts_meta.qc))

## Summary of remaining total read counts
summary(colSums(exprs(TRL_counts_meta.qc)))

(reps_per_treatment<-ddply(data.frame(pData(TRL_counts_meta.qc)),c("BCS","ExposureTime"),summarise,n=length(BCS)))

```

## Create model matrix
This is done outside edgeR GLM as it is used by other chunks, but edgeR chunk is not evaluated
```{r }
## take counts table design created above and make edgeR object
design <-model.matrix(~0+group1,data = pData(TRL_counts_meta.qc))
```
## edgeR GLM FIT
Not evualted as it takes some time. Change to eval=T to re-compute edgeR results
```{r edgeR, eval=F}
## Estimate dispersion, calc norm facots, and fit to GLM
y<-DGEList(exprs(TRL_counts_meta.qc))
y<-estimateDisp(y,design)
y<-calcNormFactors(y)
plotBCV(y,main="BCV Plot")
TRL_glmFit <- glmFit(y, design)

save(TRL_glmFit,file = paste0(R_script_output_directory,"TRL_glmFit.RData"))
```

## Perfom edgeR LRT

```{r perform_LRT}
load(paste0(R_script_output_directory,"TRL_glmFit.RData"))
## Go through each pait-wise comparison in constarsts and compute differential expression using glmLRT
contr<-makeContrasts(
  LCGV.4="group1LCRISPATUS.4HR-group1GVAGINALIS.4HR",
  LCM.4="group1LCRISPATUS.4HR-group1MEDIA.4HR",
  LJGV.4="group1LJENSENII.4HR-group1GVAGINALIS.4HR",
  LJM.4="group1LJENSENII.4HR-group1MEDIA.4HR",
  LIGV.4="group1LINERS.4HR-group1GVAGINALIS.4HR",
  LIM.4="group1LINERS.4HR-group1MEDIA.4HR",
  LCLJ.4="group1LCRISPATUS.4HR-group1LJENSENII.4HR",
  LCLI.4="group1LCRISPATUS.4HR-group1LINERS.4HR",
  LJLI.4="group1LJENSENII.4HR-group1LINERS.4HR",
  GVM.4="group1GVAGINALIS.4HR-group1MEDIA.4HR",
  
  LCGV.13="group1LCRISPATUS.13HR-group1GVAGINALIS.13HR",
  LCM.13="group1LCRISPATUS.13HR-group1MEDIA.13HR",
  LJGV.13="group1LJENSENII.13HR-group1GVAGINALIS.13HR",
  LJM.13="group1LJENSENII.13HR-group1MEDIA.13HR",
  LIGV.13="group1LINERS.13HR-group1GVAGINALIS.13HR",
  LIM.13="group1LINERS.13HR-group1MEDIA.13HR",
  LCLJ.13="group1LCRISPATUS.13HR-group1LJENSENII.13HR",
  LCLI.13="group1LCRISPATUS.13HR-group1LINERS.13HR",
  LJLI.13="group1LJENSENII.13HR-group1LINERS.13HR",
  GVM.13="group1GVAGINALIS.13HR-group1MEDIA.13HR",
  
  LCGV.22="group1LCRISPATUS.22HR-group1GVAGINALIS.22HR",
  LCM.22="group1LCRISPATUS.22HR-group1MEDIA.22HR",
  LJGV.22="group1LJENSENII.22HR-group1GVAGINALIS.22HR",
  LJM.22="group1LJENSENII.22HR-group1MEDIA.22HR",
  LIGV.22="group1LINERS.22HR-group1GVAGINALIS.22HR",
  LIM.22="group1LINERS.22HR-group1MEDIA.22HR",
  LCLJ.22="group1LCRISPATUS.22HR-group1LJENSENII.22HR",
  LCLI.22="group1LCRISPATUS.22HR-group1LINERS.22HR",
  LJLI.22="group1LJENSENII.22HR-group1LINERS.22HR",
  GVM.22="group1GVAGINALIS.22HR-group1MEDIA.22HR",
  
  levels=design
)

desets<-list()
comparisons<-names(data.frame(contr))

#postscript(paste0(R_script_output_directory,"TRL_SmearPlots.eps"),height = 8,width = 10)

for(comparison in comparisons){
  #comparison<-"LCGV.13"
  comp<-glmLRT(TRL_glmFit,contrast = contr[,comparison]) ## DE using constrast

  de.table<-comp$table[abs(comp$table$logFC)>1 & comp$table$logCPM>1 & p.adjust(comp$table$PValue,method = "fdr")<=0.01,]
  comp$table$PValue.adj<-p.adjust(comp$table$PValue,method = "fdr")
  desets[[comparison]]<-list(detags=nrow(de.table),fulltable=comp$table)
  #plotSmear(comp,de.tags = names(de.table),main=paste0(comparison,": ",length(detags)," DE genes"))

}

```


Upload DE table to IPA, run IPA pathway analysis, then save the pathway comparisons to file
Performed March-April 2017. See thesis for version/db builds. 

## Read in IPA Results

```{r IPA, eval=T}
##Timecourse files are from IPA- contain pathway list and activation z values.

pathway_zscore_files<-list.files(path = R_script_input_directory, pattern = "timecourse.txt")
pathway_zscores<-data.frame(Canonical.Pathway="DROP")
functions<-data.frame(Comparison="DROP",Categories="DROP",Diseases.or.Functions.Annotation="DROP",p.Value=1,Predicted.Activation.State="DROP",Activation.z.score=0,Flags="DROP", Molecules="DROP")

for(tab in pathway_zscore_files){
  #tab<-"LCGV_LCM_timecourse.txt"
  newt<-  read.table(paste0(R_script_input_directory,tab),header = T,sep = "\t",skip = 1,na.strings = "N/A",stringsAsFactors = F) %>% dplyr::select(-X)
  pathway_zscores<-merge(newt,pathway_zscores,all = T)
}

## Put pathway z scores into matricies and then split absed on G. vag or medium reference
pathway_zscores.matrix<-as.matrix(pathway_zscores[,2:ncol(pathway_zscores)])
row.names(pathway_zscores.matrix)<-pathway_zscores$Canonical.Pathway
pathway_zscores.matrix<-pathway_zscores.matrix[rowSums(abs(pathway_zscores.matrix),na.rm = T)>0,colSums(abs(pathway_zscores.matrix),na.rm = T)>0]

pathway_zscores.melt<-melt(pathway_zscores,id.vars = "Canonical.Pathway")
row.names(pathway_zscores)<-pathway_zscores$Canonical.Pathway
pathway_zscores.matrix[is.na(pathway_zscores.matrix)]<-0

## Create a design table for the pathways
Pathways_design<-data.frame(comparison=names(data.frame(pathway_zscores.matrix)))
Pathways_design$comp<-sapply(strsplit(as.character(Pathways_design$comparison),"\\."),function(x) x[[1]])
Pathways_design$ExposureTime<-sapply(strsplit(as.character(Pathways_design$comparison),"\\."),function(x) x[[2]])
Pathways_design$L<-sapply(strsplit(Pathways_design$comp,split = "*"),function(x) paste0(x[[1]],x[[2]]))
Pathways_design$ref<-sapply(strsplit(Pathways_design$comp,split = "*"),function(x) paste0(x[[3]]))
row.names(Pathways_design)<-Pathways_design$comparison

## Map mapthway names to classifications
path_type_map<-read.csv(paste0(R_script_input_directory,"Pathway_classification.csv"),stringsAsFactors = F)


## Subset to look at only cell culture medium references
medium_comparisons<-names(data.frame(pathway_zscores.matrix))[grepl(names(data.frame(pathway_zscores.matrix)),pattern = "M")]
pathway_zscores.matrix.medium<-pathway_zscores.matrix[,names(data.frame(pathway_zscores.matrix)) %in% medium_comparisons]

## Summary table for pathway classification
summary_pathways<-cbind(data.frame(num_cycle.p=colSums(pathway_zscores.matrix.medium[row.names(data.frame(pathway_zscores.matrix.medium)) %in% path_type_map[path_type_map$class=="c","pathway"],]>2)),
                        data.frame(num_cycle.n=colSums(pathway_zscores.matrix.medium[row.names(data.frame(pathway_zscores.matrix.medium)) %in% path_type_map[path_type_map$class=="c","pathway"],]<(-2))),
                        data.frame(num_immune.p=colSums(pathway_zscores.matrix.medium[row.names(data.frame(pathway_zscores.matrix.medium)) %in% path_type_map[path_type_map$class=="i","pathway"],]>2)),            
                        data.frame(num_immune.n=colSums(pathway_zscores.matrix.medium[row.names(data.frame(pathway_zscores.matrix.medium)) %in% path_type_map[path_type_map$class=="i","pathway"],]<(-2))),

                        data.frame(num_immune_pro.n=colSums(pathway_zscores.matrix.medium[row.names(data.frame(pathway_zscores.matrix.medium)) %in% path_type_map[path_type_map$X=="pro","pathway"],]<(-2))),

                        data.frame(num_immune_pro.p=colSums(pathway_zscores.matrix.medium[row.names(data.frame(pathway_zscores.matrix.medium)) %in% path_type_map[path_type_map$X=="pro","pathway"],]>2)))

summary_pathways$comparison<-row.names(summary_pathways)
summary_pathways<-merge(summary_pathways,Pathways_design)

## Write table summarizing the number of pathways above or below z score, grouped by pathway category
write.csv(summary_pathways,paste0(thesis_tables_directory,TABLE_TRL_SUMMARY_PATHWAYS),row.names=F,quote=F)

write.csv(pathway_zscores.matrix.medium,paste0(thesis_tables_directory,TABLE_TRL_PATHWAY_Z_SCORES),row.names=T,quote=F)

##Sort # of z>2 pathways by negative cycle, positive immune and negative immune
summary_pathways[order(summary_pathways$num_cycle.n,decreasing = T),]
summary_pathways[order(summary_pathways$num_immune.p,decreasing = T),]
summary_pathways[order(summary_pathways$num_immune.n,decreasing = T),]

## Use pathways that are expressed abs(z-score)>2 in 10% of comparisons 
pathway_zscores.matrix.medium<-pathway_zscores.matrix.medium[(rowSums(abs(pathway_zscores.matrix.medium)>2))>=.1*ncol(pathway_zscores.matrix.medium),]

## Clean up some of the pathway names in pathway map
path_type_map[path_type_map$pathway=="NF-_B Signaling","pathway"]<-"NF-B Signaling"
path_type_map[path_type_map$pathway=="PKC_ Signaling in T Lymphocytes","pathway"]<-"PKC Signaling in T Lymphocytes"

path_type_map[path_type_map$X=="pro","class"]<-"pro"

## Add in pathway classification based on mapping file
pathway_zscores.matrix.medium<-data.frame(pathway_zscores.matrix.medium)
pathway_zscores.matrix.medium$pathway<-row.names(pathway_zscores.matrix.medium)
pathway_zscores.matrix.medium<-merge(pathway_zscores.matrix.medium,path_type_map) ## maps classifications to pathway names
row.names(pathway_zscores.matrix.medium)<-pathway_zscores.matrix.medium$pathway

table(path_type_map[path_type_map$pathway %in% row.names(data.frame(pathway_zscores.matrix.medium)),"class"])

## Melt the pathway z-scores table
pathway_zscores.medium.melt<-melt(pathway_zscores.matrix.medium,id.vars = c("class","X","pathway"))
pathway_zscores.medium.melt<-dplyr::rename(pathway_zscores.medium.melt,"comparison"=variable)
pathway_zscores.medium.melt<-merge(pathway_zscores.medium.melt,Pathways_design)

## Ensure the exposure times and BCS are ordered
summary_pathways$ExposureTime<-as.numeric(summary_pathways$ExposureTime)
summary_pathways$ExposureTime<-factor(summary_pathways$ExposureTime,levels = c(4,13,22),ordered = T)
pathway_zscores.medium.melt$comp<-factor(pathway_zscores.medium.melt$comp,levels = c("LCM","LJM","LIM","GVM"),ordered = T)
pathway_zscores.medium.melt$ExposureTime<-factor(pathway_zscores.medium.melt$ExposureTime,levels = c("4","13","22"),ordered = T)
pathway_zscores.medium.melt$x<-paste(pathway_zscores.medium.melt$comp,pathway_zscores.medium.melt$pathway)

##Clean up the pathway names for better plotting
#pathway_zscores.medium.melt[pathway_zscores.medium.melt$pathway=="Role of IL-17F in Allergic Inflammatory Airway Diseases","pathway"]<-"IL-17F in Allgc. Inflam. Arwy Dis."
#pathway_zscores.medium.melt[pathway_zscores.medium.melt$pathway=="Role of Pattern Recognition Receptors in Recognition of Bacteria and Viruses","pathway"]<-"PRRs/ Bacteria and Viruses"
#pathway_zscores.medium.melt[pathway_zscores.medium.melt$pathway=="Production of Nitric Oxide and Reactive Oxygen Species in Macrophages","pathway"]<-"Production of NO and ROS in Macrophages"
#pathway_zscores.medium.melt[pathway_zscores.medium.melt$pathway=="PKC Signaling in T Lymphocytes","pathway"]<-"PKC Signaling"
#pathway_zscores.medium.melt[pathway_zscores.medium.melt$pathway=="PI3K Signaling in B Lymphocytes","pathway"]<-"PI3K Signaling"

##Subset z-scores table by cell cycle pathways
pathway_zscores.medium.melt.cycle<-dplyr::filter(pathway_zscores.medium.melt,class %in% c("c"))

unique(dplyr::select(pathway_zscores.medium.melt,c(class,pathway)))
paste0("Number of pathways belonging to each class:")
table(unique(dplyr::select(pathway_zscores.medium.melt,c(class,pathway))) %>% dplyr::select(class))

paste0(c("the remaining 5 pathways did not belong to either cell cycle or immunity:",str_c(unique(dplyr::select(pathway_zscores.medium.melt,c(class,pathway))) %>% dplyr::filter(class=="o") %>% dplyr::select(pathway),collapse = ", ")))

ggplot(pathway_zscores.medium.melt.immune)+geom_tile(aes(x=ExposureTime,y=pathway,fill=value))+
scale_fill_gradient2(high="red",mid="white",low="blue", 
       na.value="yellow", midpoint=0)+facet_wrap(~comp,nrow=1)+
  #journalFormat+
  theme_bw() + theme(text = element_text(colour = "black",size=12))
  #theme(text = element_text(size=12),plot.margin = unit(c(0,30,0,30),units = "pt"))

dev.off()

cairo_ps(paste0(thesis_figures_directory,FIGURE_COMBINED_PATHWAYS_CYCLE,".eps"),width=10,height=8)

ggplot(pathway_zscores.medium.melt.cycle)+geom_tile(aes(x=ExposureTime,y=pathway,fill=value))+
scale_fill_gradient2(high="red",mid="white",low="blue", 
       na.value="yellow", midpoint=0)+facet_wrap(~comp,nrow=1)+
  journalFormat+
  theme(text = element_text(size=12),plot.margin = unit(c(0,30,0,0),units = "pt"))

dev.off()

```

## Extract and plot logFC values from LRT table

```{r plot_FC}
## Pull out DE tables from each comparison within list, and write this table to file
edgeR_results<-data.frame(gene=row.names(exprs(TRL_counts_meta.qc)))
row.names(edgeR_results)<-edgeR_results$gene

num_de_genes<-data.frame(num_de_genes=sapply(desets,function(x) x[[1]])) 
num_de_genes$comparison<-row.names(num_de_genes)
write.csv(num_de_genes[num_de_genes$comparison %in% medium_comparisons,],file=paste0(m=thesis_tables_directory,TABLE_TRL_NUMDEGENES))

##Exctract genes from the list
for(i in names(desets)){
  #i<-"LCGV.4"
  tmp.df<-data.frame(desets[[i]]$fulltable)
  names(tmp.df)<-paste0(i,".",names(tmp.df))
  tmp.df$gene<-row.names(desets[[i]]$fulltable)
  edgeR_results<-join(edgeR_results,tmp.df,"gene")
}

hist(as.matrix(dplyr::select(edgeR_results,ends_with("logFC"))))
hist(as.matrix(dplyr::select(edgeR_results,ends_with("logCPM"))))
quantile(as.matrix(dplyr::select(edgeR_results,ends_with("logFC"))),probs = c(0.68,0.95))
edgeR_results[1:5,1:6]

row.names(edgeR_results)<-edgeR_results$gene
dplyr::select(edgeR_results[edgeR_results$gene %in% c("EGFR","EP300","HDAC4","CDKN1A"),],contains("M")) %>% dplyr::select(contains("PValue.adj"))

write.csv(edgeR_results,paste0(thesis_tables_directory,TABLE_EDGER_RESULTS),quote = F,row.names = F)
#write.table(edgeR_results,paste0(thesis_tables_directory,"TABLE_A11.txt"),quote = F,row.names = F,sep="\t")
#write.csv(dplyr::select(edgeR_results,c(gene,ends_with("logFC"),ends_with("PValue"))),paste0(root_directory,"edgeR_results_LCPval.csv"),quote = F,row.names = F)

## Clean up the table containg the DE expresison information
edgeR_results.melt<-melt(edgeR_results)
edgeR_results.melt$variable<-gsub(pattern = "PValue.adj",replacement = "Pvalue_adj",x = edgeR_results.melt$variable)
edgeR_results.melt<-separate(edgeR_results.melt,"variable",sep="\\.",into = c("comparison","ExposureTime","value_type"))

## Assign colors to BCS
color_map<-c("LCGV"=unname(subject_long_taxa_colors["Lactobacillus_crispatus"]),"LCM"=unname(subject_long_taxa_colors["Lactobacillus_crispatus"]),"LJGV"=unname(subject_long_taxa_colors["Lactobacillus_jensenii"]),"LJM"=unname(subject_long_taxa_colors["Lactobacillus_jensenii"]),"LIM"=unname(subject_long_taxa_colors["Lactobacillus_iners"]),"LIGV"=unname(subject_long_taxa_colors["Lactobacillus_iners"]),"GVM"=unname(subject_long_taxa_colors["Gardnerella_vaginalis"]))

## Use different line types for G. vaginalis and medium (if used)
line_map<-data.frame(comparison=names(color_map),line_type=c("solid","dashed","solid","dashed","dashed","solid","dashed"))

## Merge tables together to map DE results to plot annotations
edgeR_results_colors<-data.frame(comparison=names(color_map),colr=unname(color_map))
edgeR_results.melt<-merge(edgeR_results.melt,edgeR_results_colors)
edgeR_results.melt<-merge(edgeR_results.melt,line_map)
edgeR_results.melt<-merge(edgeR_results.melt,dplyr::select(Pathways_design,c(comp,L,ref)),all.x=T,by.x="comparison",by.y="comp")
edgeR_results.melt$ExposureTime<-as.numeric(edgeR_results.melt$ExposureTime)


## Format and plot the cell cycle pathway-related gene expression's logFC over the timecourse
cycle_genes<-c("HDAC4","EP300","CDKN1A","CDK4","CCND1","CCNE2","ESR1","EGFR") ## Select which cell cycle genes to plot
cycle_genes_expression<- dplyr::filter(edgeR_results.melt,gene %in% cycle_genes &  ref=="M" ) ## Only include the cell cycle genes and cell culture medium as the reference
cycle_genes_expression<-unique(cycle_genes_expression)
cycle_genes_expression<-spread(cycle_genes_expression,key = value_type,value = value) ## This will make the logFC, FDR, and other DE attributes into colums for easier plotting. The plot will use logFC and FDR information
cycle_genes_expression$gene<-factor(cycle_genes_expression$gene,levels = cycle_genes,ordered = T) ## Maintain order of genes- this follows logical order discussed in thesis. 
cycle_genes_expression[cycle_genes_expression$Pvalue_adj<0.01,"DE.pval"]<-"*" ## Annotate which samples are DE by FDR

## Make plot
long_plot.cc<-ggplot(cycle_genes_expression)+geom_point(aes(x=as.numeric(ExposureTime),y=logFC,col=comparison),size=3)+
  geom_line(aes(x=as.numeric(ExposureTime),y=logFC,col=comparison),size=2)+
  facet_wrap(~gene,scales = "free_y",nrow =2)+theme_bw()+scale_color_manual(values = color_map)+
  geom_hline(yintercept =0)+geom_hline(yintercept =c(-1,0,1),lty=2)+xlab("Exposure Time (hours)")+
  geom_text(aes(x=as.numeric(ExposureTime),y=logFC,label=DE.pval),size=8)+
  journalFormat+
  ylab("log (Fold Change)")+scale_x_continuous(breaks=c(4,13,22))

plot(long_plot.cc)

cairo_ps(paste0(thesis_figures_directory,FIGURE_LONGITDUINAL_GENEEXP.cycle),width = 8,height = 6)
plot(long_plot.cc)
dev.off()

```


# End Timestamp
```{r Timestamp2}
## Log session info
filewritable_time<-gsub(gsub(Sys.time(),pattern = " ",replacement = "_"),pattern = "-|:",replacement = "")
(sessionInfo_latex<-toLatex(sessionInfo()))
write(sessionInfo_latex,paste0(R_script_output_directory,"R_sessionInfo_",filewritable_time,".log.txt"))

timestamp()


```
